---
title: "Beyond Bergmann and Allen: A Unified Framework for Thermoregulatory Morphology"
number-sections: true
# If you do not want code to show... echo: false
execute: 
 echo: false
 message: false
 warning: false 
format: pdf
editor: visual
bibliography: ../Supp_files/EWPW.bib
link-citations: true
csl: ../Supp_files/methods-in-ecology-and-evolution.csl
knitr:
  opts_chunk: 
    comment: "#>"
    collapse: true
editor_options: 
  chunk_output_type: console
---

{{< pagebreak >}}

# To do

-Read Frohlich & Baldwin (Weeks, 2020?) more carefully. Why do we need the thermo-morpho rule?

-Recreating Frohlich Fig 4f

-Secondary: Identify body size metrics used in studies of Bergmann's rule (see Lit review text section). If all studies use multiple metrics, most of which are appendages, what's different between Allen's and Bergmann's? Accounting for allometry.

-Remember, original article idea was importance of examining multiple traits in Bergmann's rule. Have simulation code to illustrate this, which could go into boxes for GEB perspective.

-   Even Bergmann's rule benefits from a multi-trait approach, as well as examining trait covariance and allometric relationships (practices usually associated with assessment of Allen's rule). Using simulated data I show that analyzing traits in isolation can obscure ecologically and evolutionarily meaningful patterns.

-Importance of allometry? Does this make sense for this article? Include the distinct approaches in the allometry box? Frohlich article seems to think so, see example on page 3.

-Allometry is then accounted for in the final model proposed by Frohlich, which is a nice way to tie things together

-Allen himself in original article suggested interactions between body size & temperature, and supposedly framed his research in the context of Bergmann's rule

-Macedo et al (2025) integration of Bergmann's, Allen's, and Bogert's rule in a single path analysis

-Analyzing ecogeographical rules in isolation 1) slows scientific progress, 2) miss relationships (Baldwin, 89 / 107 families), and 3) obtain an incomplete or entirely misinformed understanding of wildlife response to temperature gradients (extreme variation in slopes depending on size and shape).

# Abstract

**Background:** Understanding how organismal size and shape vary across environmental gradients is essential for ecology, evolution, and conservation. Yet size and shape are often studied separately, despite their shared functional basis. Of 47 studies published since 2021 examining Bergmann's rule, Allen's rule, or both, nearly half examined Bergmann's or Allen's rule in isolation. A focus on Bergmann's rule in isolation was much more common than focusing on Allen's rule in isolation, likely because most authors (\~90%) interpreted Allen's rule as a question of appendage length *relative* to body size, making it inherently multivariate.

**A unified framework:** I propose retiring the siloed treatment of Bergmann’s and Allen’s rules in favor of a unified framework: the ***Thermo-Morpho Rule***, which views size and shape as complementary axes of thermoregulatory adaptation. These "rules" stem from the same biophysical principle - that surface area to volume ratios mediate heat exchange. Recent empirical evidence from joint analyses of Bergmann’s and Allen’s rules supports the unified framework proposed by the Thermo-Morpho Rule. These studies point to a functional tradeoff, where 1) species often respond to warming either by decreasing in body size or by increasing relative appendage length, but rarely both; and that species 2) exhibited small, concurrent changes in both traits that were difficult to detect in isolation, but still amounted to an important change in surface area to volume ratios.

**Conclusions:** Studying size and shape in tandem offers a more biologically meaningful understanding of how morphology tracks climate. In the Anthropocene, where rapid environmental change demands fast and informed decisions, conservationists can’t afford to split hairs over legacy rules. A unified approach - grounded in theory, guided by trait covariance, and aware of allometry - is overdue.

# Introduction

We are in the Anthropocene — a period of rapid global change dominated by habitat loss and climate change. While many species have not kept pace with these changes and have gone extinct, other species have been able to alter their shapes, sizes, ranges, and behaviors to adjust to this new human-dominated world. Understanding these sub-lethal, plastic or evolutionary responses is critical to understanding humanities’ full impact on biodiversity [@guoApplyingBergmannsRule2025]. With this knowledge, we will be better able to forecast changes to the natural world, as well as their impacts on ecosystem service provisioning.

An organism’s morphology is perhaps the most fundamental determinant of how it interacts with its environment, and thus there are strong selective pressures to ensure that an organism's morphology is well-suited for its environment [@pigotMacroevolutionaryConvergenceConnects2020]. Thus, scientists have long sought to understand how morphology is shaped by the environment [e.g., @winklerEcomorphologyMigrants1992], and some of the oldest macrogeographical 'rules' attempt to make sense of patterns in body size and shape. For example, Bergmann’s rule posits that individuals or populations of endothermic species tend to be larger in colder climates, presumably to minimize heat loss via a reduced surface-area-to-volume ratio [@bergmann1848ueber]. The complementary Allen’s rule states that appendages (e.g., limbs, tails, bills) tend to be shorter in colder regions, again reducing heat loss via a decreased surface-area-to-volume ratio [@allen1877influence].

Importantly, both of these rules stem from a central tenet - that changing body plan (i.e., size or shape) alters an organism's surface-area-to-volume ratio, which has implications for thermoregulation. Given their shared thermoregulatory underpinning, I propose a combined ecogeographical rule, the *Thermo-Morpho Rule*, which views size and shape as complementary axes of thermoregulatory adaptation. For scientists, the thermo-morpho rule implies that Bergmann's and Allen's rule should be examined together, given that similar changes in surface-area-to-volume could be achieved through changes in size, shape, or both. In this opinion piece, I review the literature published since 2021 to examine the extent which Bergmann's and Allen's rules have been examined in concert vs in isolation, and make a case for why their joint analysis is essential in the Anthropocene.

# A review of Allen and Bergmann

There is general supp fdsfort for both Bergmann’s and Allen's rules in endotherms - across both space and time, and among diverse taxa, regions, and life-history strategies such as migration and hibernation [@teplitskyClimateWarmingBergmanns2014; @rydingShapeshiftingChangingAnimal2021; @frohlichAllometryRevealsTradeoffs2023]. Bergmann's rule is perhaps the most extensively studied ecogeographic rule to date, and several important reviews cover adherence to the rule [and lack thereof , @ashtonPatternsWithinSpeciesBody2002], potential mechanisms [@yom-tovRecentSpatialTemporal2011], implications in the face of climate change [@gardnerDecliningBodySize2011; @teplitskyClimateWarmingBergmanns2014], and best practices for empirical studies [@theriotHarnessingNaturalHistory2023]. Allen's rule is less studied [but see @rydingShapeshiftingChangingAnimal2021], despite evidence that it is equally important as Bergmann's rule for thermoregulation [@frohlichAllometryRevealsTradeoffs2023].

While both legacy rules arise from a similar theoretical concept, the analytical approaches used to analyze the two rules has differed substantially. Generally, Bergmann's rule is assessed using one or more regression models with a metric of body size (e.g., body mass) as the response variable in each model [e.g., @skinnerEnvironmentalGeographicConditions2025]. It is recommended to include multiple morphological metrics as complementary (though non-independent) representations of body size [@theriotHarnessingNaturalHistory2023]. While this may result in conflicting signals across traits [e.g., @baileyUsingDifferentBody2020], this interpretative complexity more accurately reflects the true nature of environment–morphology relationships.

Allen's rule, on the other hand, has generally been interpreted as how appendage length changes *relative* to body size across environmental gradients [e.g., @rydingMultipleRegressionNot2025]. In other words, estimating relative appendage length requires tying appendage length and body size together *at the level of the individual* through some sort of allometric scaling relationship (@nte-allometry). This can be done in a multiple regression framework [@rydingMultipleRegressionNot2025], or by adhering to a power law definition [as recommended by @santoroAllometryEvaluateAllens2022]. Thus, Allen's rule is inherently a multivariate endeavor while Bergmann's rule could theoretically be examined using a single body size.

::: {#nte-allometry .callout-note}
## Allometric scaling in studies assessing relative appendage length

Allometry is the study of how biological processes change with body size. Seminal work by @huxley1932problems formulated that the scaling relationship between an appendage $Y$ and body size $X$ follows a power-law, where

$$
Y = \alpha * X^\beta
$$

On the log scale this function becomes additive

$$
log(Y) = log(\alpha) + \beta * log(X)\
$$

where $log(\alpha)$ determines the average relative appendage length for a given body size (the allometric intercept), and $\beta$ determines the rate that the appendage increases with an increase in body size [the allometric slope, @shingletonAllometryStudyBiological2010].

Allometric scaling theory posits that proportional growth of two linear features should increase linearly ($\beta$ = 1), whereas proportional growth of a linear feature and volume should scale inverse cubically [$\beta = 1/3$ i.e. given that organismal volume is three-dimensional, @raynerFormFunctionAvian1988]. When empirical reality matches theoretical expectations, this is known as isometric scaling [@shingletonAllometryStudyBiological2010]. If the allometric slope exceeds that expected by theory (e.g. $\beta > 1/3$), this is known as hyperallometry (or positive allometry), and when it is is less than expected by theory (e.g. $\beta < 1/3$), this is known as hypoallometry. Importantly, isometric scaling implies that body *shape* remains constant as body size increases, whereas hyper- and hypo- allometry suggest changes in body proportions.
:::

# Literature review

## Figure

```{r}
#| label: Setup

# Load libraries
library(tidyverse)
library(readxl)
library(janitor)
library(smatr)
library(ggpubr)
library(cowplot)
ggplot2::theme_set(theme_cowplot())

# Source functions
source("Scripts/Key_allometry_fns.R")
```

```{r}
#| label: Load-in-format-data

Lr_allens <- read_excel("Lit_review/Methods_ecogeographic_rules_8.12.25.xlsx", 
                         sheet = "Updated Excel") %>% 
  clean_names() %>% 
  filter(include_exclude != "Exclude") %>%
  dplyr::select(authors, year, journal, bergmann) %>% 
  mutate(focus = ifelse(bergmann == "Y", "Both", "Allens_only")) %>% 
  dplyr::select(-bergmann)
Lr_bergs <- read_excel("Lit_review/Methods_ecogeographic_rules_8.12.25.xlsx", 
                         sheet = "Bergs studies") %>% 
  clean_names() %>% 
  dplyr::select(authors, year, journal, allens) %>% 
  mutate(focus = ifelse(allens == "Y", "Both", "Bergmanns_only")) %>% 
  dplyr::select(-allens)

Lr_ecogeo <- bind_rows(Lr_allens, Lr_bergs)

# Prep labeling tbl
Lr_p <- Lr_ecogeo %>% 
  summarize(num_studies = n(), .by = c(focus)) %>% 
  mutate(Tot_studies =  sum(num_studies), 
         Prop = num_studies / Tot_studies,
         Per = Prop * 100) %>% 
  mutate(Per = round(Per, 0),
         Prop = round(Prop, 2), 
         Per = paste0(Per, "%"),
         focus = str_replace(focus, "_", "\n"))
```

```{r}
#| label: fig-barplot-in-isolation
#| fig-cap: Of 47 studies published since 2021 examining Bergmann's rule, Allen's rule, or both, nearly half examined Bergmann's or Allen's rule in isolation. A focus on Bergmann's rule in isolation was much more common than focusing on Allen's rule in isolation, likely because Allen's rule is inherently multivariate. 
#| fig-height: 6.5

#Plot
Lr_ecogeo %>% 
  mutate(focus = str_replace(focus, "_", "\n")) %>% 
  ggplot() + 
  geom_bar(aes(x = focus, fill = factor(year)), 
           color = "black", stat = "count") + 
   geom_text(
    data = Lr_p, aes(x = focus, y = num_studies, label = Per),
    vjust = -0.7, size = 5
  ) + labs(x = NULL, y = "Frequency", fill = "Year")
```

## Text

```{r}
#| label: pull-percents

pull_per <- function(focus){
  Lr_p %>% filter(focus == {{ focus }}) %>% 
  pull(Per)
}
Per_allen <- pull_per(focus = "Allens_only")
Per_berg <- pull_per(focus = "Bergmanns_only")
Per_both <- pull_per(focus = "Both")
```

Of `r nrow(Lr_ecogeo)` studies published since 2021 examining Bergmann's rule, Allen's rule, or both, nearly half examined them in isolation. A focus on Bergmann's rule alone (`r Per_berg`) was far more common than on Allen's rule alone (`r Per_allen`), likely because Allen's rule studies generally measured *relative* appendage length (89% of studies), i.e., appendage length controlling for body size. Thus, at some level studies of Allen's rule are already examining trends in body size across ecological gradients. The studies using metrics of *relative* appendage length, either implicitly or explicitly, acknowledge that allometric scaling is critical in the examination of Allen's rule [@santoroAllometryEvaluateAllens2022; @rydingMultipleRegressionNot2025].

OPTION: The majority of studies examining Bergmann's rule in isolation used multiple traits of body size. Examination of trait covariance (i.e., allometric relationships) was done in \~10% of studies. IF this is true, could then speculate why this difference arose. Bergmann's rule researchers are thinking about body size - be it approximated with appendage, mass, total length, whereas Allen's rule researchers are explicitly examining trends in body size across ecological gradients (in the context of examining relative appendage length).

Note that analyzing absolute appendage is not necessarily WRONG.

# The need for a combined approach

What do we risk by examining Bergmann's or Allen's rule in isolation? Two recent global analyses of Bergammn's and Allen's rules highlight distinct reasons why it is critical to jointly examine size and shape.

Complementarity: @baldwinComplementarityAllensBergmanns2023 found that changes in body size and relative appendage length were subtle and difficult to detect statistically. Instead, bird families exhibited small, concurrent changes in both traits in the directions predicted by Bergmann's and Allen's rules.

The 'Trade-off hypothesis': @frohlichAllometryRevealsTradeoffs2023 show that species often respond to warming either by decreasing in body size or by increasing relative appendage length, but rarely both.

Both of these studies help explain why past studies have found mixed support for Bergmann's and Allen's rules.

-   Complementarity: @baldwinComplementarityAllensBergmanns2023 found that changes in body size and relative appendage length were subtle and difficult to detect statistically. Instead, bird families exhibited small changes in both body size and appendage lengths in the directions predicted by Bergmann's and Allen's rules [@baldwinComplementarityAllensBergmanns2023]. More subtle changes in body plan could reduce the negative fitness consequences of extreme changes in either body size [e.g., a reduction in competitive ability @shipleySelectionCounteractsDevelopmental2022], or appendage length (e.g., foraging efficiency CITE van gils, acoustic signalling CITE Sebastianelli)

    -   Draw from figs 1 & 2

'Trade-off hypothesis': Similarly, @frohlichAllometryRevealsTradeoffs2023 drawing from evolutionary compromise theory found evidence for the 'trade-off hypothesis', that 'warmer temperatures tends to be associated either with smaller body size (Bergmann’s rule) or longer beak (Allen’s rule), rather than both rules simultaneously'. Thus, evolution of body size and appendage length are 'interacting ways to cope with thermoregulation', which ‘may explain why many studies fail to detect Allen’s or Bergmann’s rules independently, which has led to questioning of the generality of these ecogeographical patterns’.

-   A dizzying array of interactions (all well thought out and justified). Models with interactions were so important that despite overall conformity to both Bergmann's rule, the patterns were reversed in birds with certain body shapes. These interactions would be missed in isolation. Similarly with Allen's rule, the overall effect of temperature on relative appendage length is negative, opposite of what Allen would have predicted. However, when including the interaction between tempearture and body size it becomes clear that the only small birds exhibit inverse Allen's rule, whereas large birds do indeed adhere to Allen's rule. This nuance is important given that it more completely tests Allen's predictions, can help generate new hypotheses ('why do certain sizes and shapes support the tradeoff hypothesis while others do not'), etc.

-   Relative beak and tarsus length show independent evolutionary patterns across temperature gradients , thus need multiple metrics

-   In predicting a species experienced temperature within its geographic range, the model with a three-way interaction between body size, tarsus, and beak was the most supported, Figure 4f is a good illustration of this. Switching so a species experienced temperature within its geographic range was DV is clever way to test for three-way interactions

# The Thermo-Morpho Rule

The Thermo-Morpho Rule, which views size and shape as complementary axes of thermoregulatory adaptation, provides theoretical justification for analyzing Bergmann's and Allen's rules in a unified framework. Irrespective of the taxonomic level of analysis, adherence to the Thermo-Morpho Rule is defined by adherence to Bergmann's rule, Allen’s rule, or showing coordinated changes in both body size and relative appendage length that together amount to an important change in the surface area to volume ratio. Recent empirical evidence supports the unified framework proposed by the Thermo-Morpho Rule.

# Looking forward

## An integrated model

Moving forward, I recommend an integrated model to jointly estimate Bergmann’s and Allen’s rule, similar to that of @frohlichAllometryRevealsTradeoffs2023. These authors compared four alternative hypotheses using causal path analysis and found the strongest support for a model incorporating allometry alongside temperature's effects on body size (Bergmann's rule) and appendage length (Allen's rule).

An integrated model to jointly estimate Bergmann’s and Allen’s rule, similar to that of @frohlichAllometryRevealsTradeoffs2023. They compared 4 hypotheses using causal path analysis:

1\) Null allometric model: Mass' influence on absolute appendage length

2\) Bergmann’s rule model: The effects included in the null allometric model + temperature's influence on mass

3\) Allen’s rule model: The effects included in the null allometric model + temperature's influence on appendage length

4\) ‘Mixed’ model: The combination of models 1-3 (Allometry + Bergmann's + Allen's rules)

Interestingly, they found that the Bergmann's and Allen's rules models were about equally supported, suggesting that shifts in size and shape are equally important for thermoregulation in birds.

## A 'thermo-pheno' rule

There are several ecogeographical rules that could influence an organism's response to temperature changes (Tian & Benton 2020). Coloration is another axis of thermoregulatory adaptation and may act in complementary or contradictory ways alongside Bergmann's and Allen's rules. Bogert's rule, or the “thermal melanism hypothesis” (Clusella-Trullas et al. 2007), proposes that organisms tend to be darker in cooler environments because dark integuments absorb heat more efficiently. Although originally formulated for lizards (Bogert 1949) and broadly supported in ectotherms (Clusella-Trullas et al. 2008, 2009; Martínez-Freiría et al. 2020), recent evidence shows parallels in endotherms as well (Delhey et al. 2018; Amar et al. 2019; Ciurej et al. 2019; Rogalla et al. 2022). Despite this, coloration is rarely incorporated into studies of environment–morphology relationships, even though it may be an important component of thermoregulation, and size and shape.

The only analysis to date that jointly examined Bergmann's, Allen's, and Bogert's rules found evidence for an integrated response to thermoregulatory pressures, including a complex web of direct and indirect effects of temperature on body and appendage size, as well as coloration of non-feathered appendages across furnariids (family Furnariidae; Macedo et al. 2025). While Macedo et al. (2025) did not include interaction terms, and thus could not explicitly test for complementarity or trade-offs, the potential for such dynamics seems high and deserves greater attention. Given that temperature could drive additional changes to physiology (e.g., Hesse's rule; Hesse 1924, Müller et al 2014), the thermo-morpho rule could be expanded to the **'thermo-pheno' rule,** encompassing the integrated effects of temperature on phenotype more broadly. The integration of these additional ecogeographical rules will require additional data that is not regularly collected, but as has been the case with morphological trait databases (CITE AVONET), these data may be available soon (e.g. see [Project Plumage](https://www.zooniverse.org/projects/ghthomas/project-plumage)).

# Outstanding Questions

-   We still lack a clear understanding of how variation in size, shape, or allometric relationships translates into differences in thermoregulatory capacity [@santoroAllometryEvaluateAllens2022; but see @tattersallHeatExchangeToucan2009]. There are both theoretical and experimental lines of evidence suggesting that the observed changes in size and shape provide near-negligible changes in surface area to volume ratios and costs of thermoregulation [@nordShrinkingBodySize2024; @tabhLimitedEvidenceThat2025].

-   Understanding of the mechanism, whether these effects arise from natural selection, or adaptive vs neutral phenotypic plasticity [@tabhLimitedEvidenceThat2025]

-   Examination of the influence of size and shape on important ecological and evolutionary outcomes [@shipleySelectionCounteractsDevelopmental2022, CITE van gils]

-   Continued work on statistical approaches that accurately account for allometric scaling, measurement error [e.g., @romano2025shrinking], and other types of error [e.g., transient fluctuations, @ponziHeritabilitySelectionResponse2018].

-   Mechanism: Causal modeling

# Simulation

```{r}
#| label: Sim-bergs_rule_conflicting

Ex_df_hypo <- pmap(tibble(b_avg_12 = .22, r_12 = .3, r_13 = -.4, r_23 = -.1), gen_data)[[1]]

# NOTE:: In paper make note that we would know this will happen due to hypoallometry

Ex_df_hypo2 <- Ex_df_hypo %>% mutate(SA_V = Append^2 / Mass) #%>% 
  #filter(Append > 0 & Mass > 0 & SA_V < 5)
```

```{r}
#| label: Confirm-sma-slope
#| include: false

smatr::sma(Append ~ Mass, data = Ex_df_hypo2)
ref_slope <- 0.33
ref_intercept <- mean(Ex_df_hypo2$Append) - ref_slope * mean(Ex_df_hypo2$Mass)

p4 <- ggplot(Ex_df_hypo2, aes(x = Mass, y = Append)) +
  geom_point(alpha = .4) + 
  ggpmisc::stat_ma_line(method = "SMA", color = "blue") +
  geom_abline(slope = 0.33, intercept = ref_intercept, 
              linetype = "dashed", color = "red")
```

```{r}
#| label: fig-bergs_rule_conflicting

p1 <- ggplot(Ex_df_hypo2, aes(x = Temp_inc, y = Append)) +
  geom_point(alpha = .4) + 
  geom_smooth(method = "lm", se = FALSE) + 
  labs(x = "Temperature increase", y = "Appendage") 

p2 <- ggplot(Ex_df_hypo2, aes(x = Temp_inc, y = Mass)) +
  geom_point(alpha = .4) + 
  geom_smooth(method = "lm", se = FALSE) + 
  labs(x = "Temperature increase", y = "Mass")

# We would expect that SA:V would increase as temperature increases, but in this case SA:V decreases
p3 <- ggplot(Ex_df_hypo2, aes(x = Temp_inc, y = SA_V)) +
  geom_point(alpha = .4) + 
  geom_smooth(method = "lm", se = FALSE) + 
  labs(y = "Surface area to volume", x = "Temperature increase")

ggarrange(p4, p1, p2, p3, nrow = 2, ncol = 2, labels = 'AUTO')
```

Thus, a In the practical study of body size, there are several linear measurements that are regularly taken when processing birds (e.g., Append length, tarsus length), but body mass is the only metric regularly taken that is proportional to volume. Studies that do use multiple metrics for body size often use one (or more) linear metric, as well as mass, to represent body size, but the relationship between the linear metric and mass is rarely examined empirically.

# Both surface area to volume

-Some articles have considered a decrease in mass while maintaining appendage length as consistent with Allen’s rule

# Extras

Some traits are highly ecologically constrained in certain species [e.g., tarsus length in Apodiformes, bill length in Trochilidae and Rallidae; @baldwinComplementarityAllensBergmanns2023]

Inclusion of allometry in studies of Bergmann's rule can help with interpretation, but is not necessary. Dolphin article, compare allometric slopes between different populations, may aid in generating hypotheses. General reporting of scaling coefficients is helpful \[Warton\], but overall this is a different question.

Because the underlying processes shaping these spatial and temporal patterns are likely similar, we do not differentiate further between them in this paper. Our recommendations apply equally to both contexts.

complementarity; @baldwinComplementarityAllensBergmanns2023, ROODENRIJS et al black oystercatchers

One critical component in the study of Bergmann's rule is the metric used to approximate body size [@teplitskyClimateWarmingBergmanns2014, @baileyUsingDifferentBody2020; @theriotHarnessingNaturalHistory2023]. Bergmann himself used body size as a proxy for an organism’s surface-area-to-volume ratio, due to the impracticality of directly measuring surface area and volume in wildlife \[but see CITE Ryding et al 2025\]. Authors have recommended both linear proxies (e.g., tarsus, wing, hindfoot) and volumetric proxies (e.g., body mass), but their validity as a proxy of body size is context- and taxa- dependent CITE. The trade-offs and ecological influences (outside of thermoregulation) of different body size measures are addressed in reviews [@yom-tovRecentSpatialTemporal2011; @theriotHarnessingNaturalHistory2023] and in the empirical literature [e.g., @goslerFieldDeterminationBody1998; @subasingheRepeatabilityValidityPhenotypic2021]. It is critical to include multiple morphological metrics as complementary (though non-independent) representations of body size [@theriotHarnessingNaturalHistory2023]. In practice, this often involves a set of regression models, each with a different morphological trait as the response variable. Although this may result in conflicting signals across traits [e.g., @baileyUsingDifferentBody2020], this interpretative complexity more accurately reflects the true nature of environment–morphology relationships. Moreover, it presents an opportunity for hypothesis generation - for example, asking why trait A might be more sensitive \[or ecologically constrained, CITE \] to temperature than trait B in a given species or region.

In this article I use the stricter power-law definition proposed by @huxley1932problems as there is ample evidence that this is appropriate for studies examining body proportions (CITE).

::: callout-note
Understanding the allometric scaling relationship for a species can aid in interpretation of results. For example, imagine a study using two variables as metrics of body size (wing chord and mass), and a third variable $\text{wing length}^2$ divided by body mass as a more direct metric of the surface area to volume ratio [e.g. @nwaoguTemperatureAridityDetermine2018]. The authors are surprised as the hypothetical species appears to adhere to Bergmann's rule when mass and wing chord are used separately as the response variables, but they find a result contradicting Bergmann's rule when using $\text{wing length}^2$ divided by body mass as the response variable.
:::

{{< pagebreak >}}

# References
