---
title: "Beyond Bergmann and Allen: A Unified Framework for Thermoregulatory Morphology"
number-sections: true
# If you do not want code to show... echo: false
execute: 
 echo: false
 message: false
 warning: false 
format: pdf
editor: visual
bibliography: ../Supp_files/EWPW.bib
link-citations: true
csl: ../Supp_files/methods-in-ecology-and-evolution.csl
knitr:
  opts_chunk: 
    comment: "#>"
    collapse: true
editor_options: 
  chunk_output_type: console
---

{{< pagebreak >}}

# To do

-Reach out to Frohlich

-Consider McQueen et al 2022 (read box)

-Secondary: Identify body size metrics used in studies of Bergmann's rule (see Lit review text section). If all studies use multiple metrics, most of which are appendages, what's different between Allen's and Bergmann's? Accounting for allometry.

-Remember, original article idea was importance of examining multiple traits in Bergmann's rule. Have simulation code to illustrate this, which could go into boxes for GEB perspective.

-   Even Bergmann's rule benefits from a multi-trait approach, as well as examining trait covariance and allometric relationships (practices usually associated with assessment of Allen's rule). Using simulated data I show that analyzing traits in isolation can obscure ecologically and evolutionarily meaningful patterns.

-Importance of allometry? Does this make sense for this article? Include the distinct approaches in the allometry box? Frohlich article seems to think so, see example on page 3.

-Allometry is then accounted for in the final model proposed by Frohlich, which is a nice way to tie things together

-Allen himself in original article suggested interactions between body size & temperature, and supposedly framed his research in the context of Bergmann's rule

# Abstract

**Background:** Understanding how organismal size and shape vary across environmental gradients is essential in ecology, evolution, and conservation science. Yet size and shape are often studied separately, despite their shared functional basis. Of 48 studies published since 2021 examining Bergmann's rule, Allen's rule, or both, nearly half examined Bergmann's or Allen's rule in isolation. A focus on Bergmann's rule in isolation was much more common than focusing on Allen's rule in isolation, likely because most authors (\~90%) interpreted Allen's rule as a question of appendage length *relative* to body size, making it inherently multivariate.

**A unified framework:** I propose retiring the siloed treatment of Bergmann’s and Allen’s rules in favor of a unified framework: the ***Thermo-Morpho Rule***. The Thermo-Morpho Rule views size and shape as complementary axes of thermoregulatory adaptation, as both "rules" stem from the same biophysical principle - that surface area to volume ratios mediate heat exchange. Recent empirical evidence from joint analyses of Bergmann’s and Allen’s rules supports the unified framework proposed by the Thermo-Morpho Rule. These studies point to a functional tradeoff, where 1) species often responded to warming either by decreasing in body size or by increasing relative appendage length, but rarely both; and that species 2) exhibited small, concurrent changes in both traits that were difficult to detect in isolation, but still amount to an important change in surface area to volume ratios.

**Conclusions:** Studying size and shape in tandem offers a more biologically meaningful understanding of how morphology tracks climate. In the Anthropocene, where rapid environmental change demands fast and informed decisions, conservationists can’t afford to split hairs over legacy rules. A unified approach - grounded in theory, guided by trait covariance, and aware of allometry - is overdue.

# Introduction

We are in the Anthropocene — a period of rapid global change dominated by habitat loss and climate change. While many species have not kept pace with these changes and have gone extinct, other species have been able to alter their shapes, sizes, ranges, and behaviors to adjust to this new human-dominated world [@parmesanEcologicalEvolutionaryResponses2006; @gardnerDecliningBodySize2011; @rydingShapeshiftingChangingAnimal2021]. Understanding these sub-lethal, plastic or evolutionary responses is critical to understanding humanities’ full impact on biodiversity [@guoApplyingBergmannsRule2025]. With this knowledge, we will be better able to forecast changes to the natural world, as well as their impacts on ecosystem service provisioning.

An organism’s morphology is perhaps the most fundamental determinant of how it interacts with its environment, and thus there are strong selective pressures to ensure that an organism's morphology is well-suited for its environment [@pigotMacroevolutionaryConvergenceConnects2020]. Thus, scientists have long sought to understand how morphology is shaped by the environment [e.g., @winklerEcomorphologyMigrants1992], and some of the oldest macrogeographical 'rules' attempt to make sense of patterns in body size and shape. For example, Bergmann’s rule posits that individuals or populations of endothermic species tend to be larger in colder climates, presumably to minimize heat loss via a reduced surface-area-to-volume ratio [@bergmann1848ueber]. The complementary Allen’s rule states that appendages (e.g., limbs, tails, bills) tend to be shorter in colder regions, again reducing heat loss via a decreased surface-area-to-volume ratio [@allen1877influence].

There is general support for both Bergmann’s and Allen's rules in endotherms [@teplitskyClimateWarmingBergmanns2014, @rydingShapeshiftingChangingAnimal2021, @meiriValidityBergmannsRule2003], although there are also several important exceptions [@fanBergmannsRuleAllens2019, @stevensonAllensRuleNorth1986, @geistBergmannsRuleInvalid1987, @cuiFunctionalAdaptationRather2020]. Bergmann's rule is perhaps the most extensively studied macroecological rule to date, and several important reviews cover adherence to the rule [and lack thereof, @ashtonPatternsWithinSpeciesBody2002], potential mechanisms [@yom-tovRecentSpatialTemporal2011], implications in the face of climate change [@gardnerDecliningBodySize2011; @teplitskyClimateWarmingBergmanns2014], and best practices for empirical studies [@theriotHarnessingNaturalHistory2023]. Allen's rule is less studied [but see @rydingShapeshiftingChangingAnimal2021], despite evidence that it is equally important for thermoregulation [@frohlichAllometryRevealsTradeoffs2023].

Importantly, both rules stem from a central tenet: changes in body size or shape alter an organism’s surface-area-to-volume ratio, with direct consequences for thermoregulation. Scaling theory shows that heat production (via metabolism) generally scales with body volume to the 3/4 power, while the area available for heat exchange scales with volume to the 2/3 power (assuming isometric growth, i.e., no change in shape). Thus, as size increases, heat production outpaces heat dissipation, unless the organism becomes proportionally longer or otherwise alters its shape (CITE).

Given this shared thermoregulatory basis. I propose a combined ecogeographical rule, the *Thermo-Morpho Rule*, which views size and shape as complementary axes of thermoregulatory adaptation. For scientists, the Thermo-Morpho rule implies that Bergmann's and Allen's rule should be examined together, given that similar changes in surface-area-to-volume could be achieved through changes in size, shape, or both (@fig-sa-to-v). In this opinion piece, I review the literature published since 2021 to how often these rules have been examined in concert versus in isolation in birds and mammals. I argue that their joint analysis is essential for understanding thermoregulatory capacity in the context of rapid climate change and habitat conversion [@guoApplyingBergmannsRule2025].

![**Complementarity in Bergmann's and Allen's rules with increasing temperatures.** In response to increasing temperatures, organisms can achieve changes to their surface area to volume ratios by altering their a) body size, b) relative appendage size, c) or through complementarity, i.e. more minute changes to both body and appendage size. In c), morphological response in both body and appendage size is less than that expected if Bergmann's and Allen's rules were operating in isolation. Depicted examples were simulated by approximating a bird’s body with two spheres and one cone (cartoon depictions were drawn to exemplify the potential subtlety of these changes). See Figure 2 in Baldwin et al (2023) for additional details. Figure reproduced with permission from Justin Baldwin.](../Figures/Baldwin_fig2.png){#fig-sa-to-v}

# Literature review

## Figure

```{r}
#| label: Setup

# Load libraries
library(tidyverse)
library(readxl)
library(janitor)
library(smatr)
library(ggpubr)
library(cowplot)
ggplot2::theme_set(theme_cowplot())

# Source functions
source("Scripts/Key_allometry_fns.R")
```

```{r}
#| label: Load-in-format-data

Lr_allens <- read_excel("Lit_review/Methods_ecogeographic_rules_8.24.25.xlsx", 
                         sheet = "Updated Excel") %>% 
  clean_names() %>% 
  filter(include_exclude != "Exclude") %>%
  dplyr::select(authors, year, journal, bergmann) %>% 
  mutate(focus = ifelse(bergmann == "Y", "Both", "Allens_only")) %>% 
  dplyr::select(-bergmann)
Lr_bergs <- read_excel("Lit_review/Methods_ecogeographic_rules_8.24.25.xlsx", 
                         sheet = "Bergs studies") %>% 
  clean_names() %>% 
  dplyr::select(authors, year, journal, allens) %>% 
  mutate(focus = ifelse(allens == "Y", "Both", "Bergmanns_only")) %>% 
  dplyr::select(-allens)

Lr_ecogeo <- bind_rows(Lr_allens, Lr_bergs)

# Prep labeling tbl
Lr_p <- Lr_ecogeo %>% 
  summarize(num_studies = n(), .by = c(focus)) %>% 
  mutate(Tot_studies =  sum(num_studies), 
         Prop = num_studies / Tot_studies,
         Per = Prop * 100) %>% 
  mutate(Per = round(Per, 0),
         Prop = round(Prop, 2), 
         Per = paste0(Per, "%"),
         focus = str_replace(focus, "_", "\n"))
```

```{r}
#| label: fig-barplot-in-isolation
#| fig-cap: Bergmann's and Allen's rules are often examined in isolation. Of 48 studies published since 2021 examining Bergmann's rule, Allen's rule, or both, nearly half examined Bergmann's or Allen's rule in isolation. A focus on Bergmann's rule in isolation was much more common than focusing on Allen's rule in isolation, likely because Allen's rule is inherently multivariate. 
#| fig-height: 6.5

#Plot
Lr_ecogeo %>% 
  mutate(focus = str_replace(focus, "_", "\n")) %>% 
  ggplot() + 
  geom_bar(aes(x = focus, fill = factor(year)), 
           color = "black", stat = "count") + 
   geom_text(
    data = Lr_p, aes(x = focus, y = num_studies, label = Per),
    vjust = -0.7, size = 5
  ) + labs(x = NULL, y = "Frequency", fill = "Year")
```

## Text

```{r}
#| label: pull-percents

pull_per <- function(focus){
  Lr_p %>% filter(focus == {{ focus }}) %>% 
  pull(Per)
}
Per_allen <- pull_per(focus = "Allens\nonly")
Per_berg <- pull_per(focus = "Bergmanns\nonly")
Per_both <- pull_per(focus = "Both")
```

Of `r nrow(Lr_ecogeo)` studies published since 2021 examining Bergmann's rule, Allen's rule, or both, nearly half examined them in isolation. A focus on Bergmann's rule alone (`r Per_berg`) was far more common than on Allen's rule alone (`r Per_allen`), likely because Allen's rule studies generally measured *relative* appendage length (89% of studies), i.e., appendage length controlling for body size. Thus, Allen's rule is inherently a multivariate endeavor, and at some level, authors are already considering trends in body size across ecological gradients.

# The need for a combined approach

What do scientists risk by examining Bergmann's or Allen's rule in isolation? Beyond a loss of conceptual clarity, we may miss important shifts in surface area–to–volume ratios or reach incomplete (or even misleading) conclusions about wildlife responses to temperature gradients. Two recent global analyses assessing Bergmann's and Allen's rules in tandem [@baldwinComplementarityAllensBergmanns2023, @frohlichAllometryRevealsTradeoffs2023] highlight important coevolution in size and shape of birds. Both studies argue that the surface area to volume ratio is the key trait under selection, and that Bergmann's and Allen's rules do not need to both operate simultaneously in order to achieve relevant changes in the surface area to volume ratio.

@baldwinComplementarityAllensBergmanns2023 show that wildlife can achieve comparable changes in their surface area to volume ratios through simultaneous changes in both body size and appendage length that are smaller than predicted by Bergmann's or Allen's rules in isolation. This notion, called complementarity, was highly supported in their study as 81% of the 107 bird families examined did not conform to either Bergmann’s or Allen’s rules when traits were analyzed independently, and no families conformed to both rules simultaneously. Instead, bird families exhibited small coordinated shifts in both body size and appendage length in the directions predicted by Bergmann’s and Allen’s rules. These more modest adjustments in body plan may buffer against the fitness costs of extreme changes in a single trait, while still allowing for important changes in the surface area to volume ratio (@fig-sa-to-v). For example, smaller body size could lead to reduced competitive ability [@shipleySelectionCounteractsDevelopmental2022] or constrain foraging efficiency [@vangilsBodyShrinkageDue2016], and similarly, appendage size and shape influence acoustic performance [@friedmanEvolutionMultifunctionalTrait2019, @sebastianelliContinentwidePatternsSong2022].

Drawing on evolutionary compromise theory, @frohlichAllometryRevealsTradeoffs2023 found that species often respond to warming either by decreasing body size or by increasing relative appendage length, but rarely both. They suggest that size and shape have coevolved in birds, and thus there are important trade-offs associated with getting smaller (Bergmann's rule) or relatively longer (Allen's rule) with warming climates. Indeed, models with extensive interactions revealed codependence between size and shape in the evolutionary struggle for thermoregulation. For example, despite overall conformity to Bergmann’s rule, patterns were reversed (i.e., inverse Bergmann's rule) in birds with particularly short bills and long legs. Similarly with Allen's rule, the overall effect of temperature on relative appendage length was negative (i.e., birds exhibited inverse Allen's rule); however, once interactions with body size were included, it became clear that this inverse pattern applied only to small-bodied birds, whereas large-bodied birds conformed to Allen’s predictions. Conclusions would have been highly obfuscated without modeling the interactions between size, shape, and temperature, highlighting the need to include interactions when examining the codependency of ecogeographical rules [@gastonEcogeographicalRulesElements2008]. Finally, when predicting the temperatures experienced across species’ ranges, the strongest models supported a three-way interaction between body size, tarsus length, and beak length (see @fig-interactions). Formulating the experienced climate envelope as the response variable was an especially clever way to reveal the multidimensional environment-morphology relationship.

![**Modeling interactions between size and shape is critical in understanding tradeoffs in organismal response to increasing temperatures**. In predicting a species' experienced climatic (temperature) envelope, Fröhlich et al (2023) found that the environment-morphology relationship of mass (top left), beak (top right), and tarsus (bottom) was dependent on the size and shape of the organism. See Figure 4 in Fröhlich et al (2023) for additional infromation; figure reproduced with permission from Arkadiusz Fröhlich.](../Figures/Frohlich_fig4f.png){#fig-interactions}

Together, these studies illustrate why joint analyses of size and shape are critical to fully understand the nuanced morphological responses to temperature gradients. A joint analysis allows for a better understanding of the correlated responses that species exhibit in the evolution of size and shape. Indeed, @baldwinComplementarityAllensBergmanns2023 showed there was a tight negative relationship between changes in body size and appendage length, with larger shifts in one trait associated with smaller shifts in the other. A joint analytical approach can also allow for a deeper understanding of the evolutionary drivers behind adherence (or lack thereof) to these rules. For example, @baldwinComplementarityAllensBergmanns2023 showed that species that adhered to Bergmann's rule had less common bill shapes, likely because bird bills are highly functionally constrained by foraging [e.g., Trochilidae and Rallidae, @baldwinComplementarityAllensBergmanns2023]. Thus, these taxa relied on changes in body size to meet thermoregulatory needs.

These studies help explain why there has often been mixed support for Bergmann's and Allen's rules when they are studied in isolation [@fanBergmannsRuleAllens2019, @stevensonAllensRuleNorth1986, @geistBergmannsRuleInvalid1987, @cuiFunctionalAdaptationRather2020]. Indeed, it is likely that most species are altering surface area to volume ratios across temperature gradients, but researchers may (by chance) be testing the trait (i.e., size or appendage) that does not respond to temperature gradients. Alternatively, certain clades surely exhibit complementarity, where organisms exhibit changes in both size and shape in the directions predicted by Bergmann's and Allen's rules, but small effect sizes can go undetected when rules are examined in isolation.

# Looking forward

## The Thermo-Morpho Rule

The Thermo-Morpho Rule, which views size and shape as complementary axes of thermoregulatory adaptation, provides theoretical justification for analyzing Bergmann's and Allen's rules in a unified framework. Irrespective of the taxonomic level of analysis, adherence to the Thermo-Morpho Rule is defined by adherence to Bergmann's rule, Allen’s rule, or showing coordinated changes in both body size and relative appendage length that together amount to an important change in the surface area to volume ratio. Recent empirical evidence supports the unified framework proposed by the Thermo-Morpho Rule.

## The 'Thermo-Pheno' rule

There are a few additional ecogeographical rules that generate predictions for organismal response to temperature changes [@goldenbergLinkBodySize2022]. For example, Hesse's rule states that an organism's relative heart (AND LUNG?) size will be larger in cooler climates than in warmer climates to meet higher metabolic demands and support blood circulation under colder conditions (CITE Hesse 1924, Müller et al 2014). Thus, it is predicted that warming climates will shrink relative heart size [@tianPredictingBioticResponses2020], which could alleviate some thermoregulatory pressure if metabolic rates also drop (CITE Seebacher, 2009). Similarly, Bogert's rule, or the “thermal melanism hypothesis” (Clusella-Trullas et al. 2007), proposes that organisms tend to be darker in cooler environments because dark integuments absorb heat more efficiently than lighter ones. Although originally formulated for ectotherms (Bogert 1949, Clusella-Trullas et al. 2008, 2009; Martínez-Freiría et al. 2020), recent evidence shows parallels in endotherms as well (Delhey et al. 2018; Amar et al. 2019; Ciurej et al. 2019; Rogalla et al. 2022). Despite its importance for thermoregulation, additional ecogeographical rules are rarely incorporated into studies of environment–morphology relationships [but see @cuiFunctionalAdaptationRather2020], even though they may act in complementary ways alongside Bergmann's and Allen's rules [@cardiliniTestsEcogeographicalRelationships2016].

For example, the only analysis to date that jointly examined Bergmann's, Allen's, and Bogert's rules found evidence for an integrated response to thermoregulatory pressures, including a complex web of direct and indirect effects of temperature on body and appendage size, as well as coloration of non-feathered appendages, across furnariids [family Furnariidae, @macedoMacroevolutionBodyExtremities2025]. While @macedoMacroevolutionBodyExtremities2025 did not include interaction terms in models, and thus could not explicitly test for complementarity or trade-offs, the potential for such dynamics seems high and deserves greater attention. In addition to coloration, temperature could also drive changes to physiology (e.g., metabolic rate; Clarke, Rothery, & Isaac, 2010), and thus it may be more appropriate to expand the Thermo-Morpho rule to the **'Thermo-Pheno' rule,** encompassing the integrated effects of temperature on phenotype more broadly. The integration of these additional ecogeographical rules will require data that are not yet routinely collected, but as has been the case with morphological trait databases [@tobiasAVONETMorphologicalEcological2022], these data will likely become more readily available in the near future (e.g. see [Project Plumage](https://www.zooniverse.org/projects/ghthomas/project-plumage)).

## An integrated model

Moving forward, I recommend integrated models that jointly estimate Bergmann’s, Allen’s, and potentially other important ecogeographical rules [e.g., @frohlichAllometryRevealsTradeoffs2023; @macedoMacroevolutionBodyExtremities2025]. For example, @frohlichAllometryRevealsTradeoffs2023 compared four hypotheses using phylogentically-controlled causal path analysis and found the strongest support for a model incorporating allometry alongside temperature's effects on body size (Bergmann's rule) and appendage length (Allen's rule).

1\) Null allometric model: Mass' influence on absolute appendage length

2\) Bergmann’s rule model: The effects included in the null allometric model + temperature's influence on mass

3\) Allen’s rule model: The effects included in the null allometric model + temperature's influence on appendage length

4\) ‘Mixed’ model: The combination of models 1-3 (Allometry + Bergmann's + Allen's rules)

Interestingly, they found that the Bergmann's and Allen's rules models were about equally supported, suggesting that shifts in size and shape are equally important for thermoregulation in birds.

# Outstanding Questions

1.  Why do most taxonomic groups adhere to either Bergmann's *or* Allen's rule, others adhere to both Bergmann's *and* Allen's rules [e.g., European bats, @paltrinieriEffectsClimateBat2025; Australian shorebirds, @rydingLongShorttermResponses2024], and rarely, some taxa adhere to *neither* rule [@cuiFunctionalAdaptationRather2020]?

2.  We still lack a clear understanding of how variation in size, shape, or allometric relationships translates into differences in thermoregulatory capacity [@santoroAllometryEvaluateAllens2022; but see @tattersallHeatExchangeToucan2009]. There are both theoretical and experimental lines of evidence suggesting that the observed changes in size and shape provide near-negligible changes in surface area to volume ratios and thermoregulatory efficiency [@nordShrinkingBodySize2024; @tabhLimitedEvidenceThat2025].

3.  What is the mechanism behind observed morphological change? Do these effects arise from natural selection, or adaptive vs neutral phenotypic plasticity [@tabhLimitedEvidenceThat2025]?

4.  Despite the evidence that these morphological shifts provide only minor changes to thermoregulatory capacity, they may have important ecological and evolutionary outcomes [@shipleySelectionCounteractsDevelopmental2022; @vangilsBodyShrinkageDue2016]. Studies incorporating survival or reproductive success into models are critical, as this could further our understanding of outstanding questions 2 and 3.

5.  The Anthropocene is categorized by climate change, habitat loss, and conversion of natural habitats to human-dominated ones. While there has been significant strides made in understanding the impact of climate change on morphology, the research on size and shape in fragmented or human-dominated landscapes lags far behind (but see CITE Greenberg et al 2012; Schmidt & Jensen 2003; Hermes et al 2016).

6.  It is critical that we continue developing statistical approaches that jointly test ecogeographical rules of interest, while robustly accounting for allometric scaling [e.g., @frohlichAllometryRevealsTradeoffs2023; @macedoMacroevolutionBodyExtremities2025], measurement error [e.g., @romano2025shrinking], and other types of error [e.g., @ponziHeritabilitySelectionResponse2018]. Studies examining the mechanisms underlying phenotypic change (e.g., through causal modeling approaches, CITE) are also important to make more accurate predictions for future scenarios in the Anthropocene. To date, no study has adequately accounted for all of these in a joint analytical modeling approach.

7.  With a better understanding of outstanding questions 1 - 6, we can apply our knowledge towards conservation action [e.g., @guoApplyingBergmannsRule2025]. But what exactly do conservation measures look like in light of our understanding of observed phenotypic changes in the Anthropocene?

# Conclusions

Substantial resources go into collecting and analyzing morphological data. In the Anthropocene, where rapid environmental change demands fast, informed decisions, conservationists cannot afford to split hairs over legacy rules. Indeed, the hyperfocus on classifying trait variation into separate rules is ultimately unhelpful [@goldenbergLinkBodySize2022]. Instead, it is more productive to analyze multiple traits simultaneously in the holistic framework to understand how entire phenotypes respond to temperature and other environmental gradients. These recommendations are especially urgent in the face of climate change, and apply across both intra- and inter- specific scales. Intraspecific analyses can forecast how species will respond to rising temperatures within their ranges and identify vulnerable populations, whereas interspecific analyses can illuminate how entire clades respond and highlight groups of organisms particularly susceptible to warming. Ultimately, identifying which organisms exhibit particular forms of phenotypic change (e.g., shifts in coloration, body or organ size, appendage length) will enable conservationists to develop informed strategies in an increasingly warm, human-dominated world. The Thermo-Morpho rule provides the theoretical framework that demands a unified approach to understanding environment-morphology relationships.

# Data availability

All studies included in the literature review are available for download at LINK, and the full list of citations are available in the Supporting Information.

# Letter to editor

Nature comms - 'avoid repeating information that is already present in the abstract and introduction'

Authors have advocated for the integration of macroecological rules across broad trait 'categories' (e.g., morphology, physiology, life history, population dynamics; Blackburn & Gaston 2001, Gaston, Chown, and Evans 2008), yet there has been no similar perspective arguing for this integration within morphological traits, and even less a review assessing the state of the literature.

This is critical and is recommended in some empirical studies, but no clear synthesis exists

Format - Review, don't need figure / to cite all the articles , could be included in Supp information

# Extras

## Simulation

```{r}
#| label: Sim-bergs_rule_conflicting

Ex_df_hypo <- pmap(tibble(b_avg_12 = .22, r_12 = .3, r_13 = -.4, r_23 = -.1), gen_data)[[1]]

# NOTE:: In paper make note that we would know this will happen due to hypoallometry

Ex_df_hypo2 <- Ex_df_hypo %>% mutate(SA_V = Append^2 / Mass) #%>% 
  #filter(Append > 0 & Mass > 0 & SA_V < 5)
```

```{r}
#| label: Confirm-sma-slope
#| include: false

smatr::sma(Append ~ Mass, data = Ex_df_hypo2)
ref_slope <- 0.33
ref_intercept <- mean(Ex_df_hypo2$Append) - ref_slope * mean(Ex_df_hypo2$Mass)

p4 <- ggplot(Ex_df_hypo2, aes(x = Mass, y = Append)) +
  geom_point(alpha = .4) + 
  ggpmisc::stat_ma_line(method = "SMA", color = "blue") +
  geom_abline(slope = 0.33, intercept = ref_intercept, 
              linetype = "dashed", color = "red")
```

```{r}
#| label: fig-bergs_rule_conflicting

p1 <- ggplot(Ex_df_hypo2, aes(x = Temp_inc, y = Append)) +
  geom_point(alpha = .4) + 
  geom_smooth(method = "lm", se = FALSE) + 
  labs(x = "Temperature increase", y = "Appendage") 

p2 <- ggplot(Ex_df_hypo2, aes(x = Temp_inc, y = Mass)) +
  geom_point(alpha = .4) + 
  geom_smooth(method = "lm", se = FALSE) + 
  labs(x = "Temperature increase", y = "Mass")

# We would expect that SA:V would increase as temperature increases, but in this case SA:V decreases
p3 <- ggplot(Ex_df_hypo2, aes(x = Temp_inc, y = SA_V)) +
  geom_point(alpha = .4) + 
  geom_smooth(method = "lm", se = FALSE) + 
  labs(y = "Surface area to volume", x = "Temperature increase")

ggarrange(p4, p1, p2, p3, nrow = 2, ncol = 2, labels = 'AUTO')
```

Understanding the allometric scaling relationship for a species can aid in interpretation of results. For example, imagine a study using two variables as metrics of body size (wing chord and mass), and a third variable $\text{wing length}^2$ divided by body mass as a more direct metric of the surface area to volume ratio [e.g. @nwaoguTemperatureAridityDetermine2018]. The authors are surprised as the hypothetical species appears to adhere to Bergmann's rule when mass and wing chord are used separately as the response variables, but they find a result contradicting Bergmann's rule when using $\text{wing length}^2$ divided by body mass as the response variable.

## Text - general 

The inclusion of interactions between temperature, size, and shape can reveal the co-evolution of thermoregulatory strategies in wildlife.

## Text - multiple traits

KEY here is that patterns can also be missed if we do not include multiple metrics of body size or appendage length, as some traits are constrained by ecological function [i.e., foraging, locomotion, acoustic signaling, @alhajeriGlobalTestAllens2020, @baldwinComplementarityAllensBergmanns2023, @friedmanEvolutionMultifunctionalTrait2019, @meiriValidityBergmannsRule2003]. 'The length of the two different appendages - beak and tarsus - show independent evolutionary patterns' [@frohlichAllometryRevealsTradeoffs2023].

Thus, a In the practical study of body size, there are several linear measurements that are regularly taken when processing birds (e.g., Appendage length, tarsus length), but body mass is the only metric regularly taken that is proportional to volume. Studies that do use multiple metrics for body size often use one (or more) linear metric, as well as mass, to represent body size, but the relationship between the linear metric and mass is rarely examined empirically.

Inclusion of allometry in studies of Bergmann's rule can help with interpretation, but is not necessary. Dolphin article, compare allometric slopes between different populations, may aid in generating hypotheses. General reporting of scaling coefficients is helpful \[Warton\], but overall this is a different question.

Because the underlying processes shaping these spatial and temporal patterns are likely similar, we do not differentiate further between them in this paper. Our recommendations apply equally to both contexts.

complementarity; @baldwinComplementarityAllensBergmanns2023, ROODENRIJS et al black oystercatchers

One critical component in the study of Bergmann's rule is the metric used to approximate body size [@teplitskyClimateWarmingBergmanns2014, @baileyUsingDifferentBody2020; @theriotHarnessingNaturalHistory2023]. Bergmann himself used body size as a proxy for an organism’s surface-area-to-volume ratio, due to the impracticality of directly measuring surface area and volume in wildlife \[but see CITE Ryding et al 2025\]. Authors have recommended both linear proxies (e.g., tarsus, wing, hindfoot) and volumetric proxies (e.g., body mass), but their validity as a proxy of body size is context- and taxa- dependent CITE. The trade-offs and ecological influences (outside of thermoregulation) of different body size measures are addressed in reviews [@yom-tovRecentSpatialTemporal2011; @theriotHarnessingNaturalHistory2023] and in the empirical literature [e.g., @goslerFieldDeterminationBody1998; @subasingheRepeatabilityValidityPhenotypic2021]. It is critical to include multiple morphological metrics as complementary (though non-independent) representations of body size [@theriotHarnessingNaturalHistory2023]. In practice, this often involves a set of regression models, each with a different morphological trait as the response variable. Although this may result in conflicting signals across traits [e.g., @baileyUsingDifferentBody2020], this interpretative complexity more accurately reflects the true nature of environment–morphology relationships. Moreover, it presents an opportunity for hypothesis generation - for example, asking why trait A might be more sensitive \[or ecologically constrained, CITE \] to temperature than trait B in a given species or region.

In this article I use the stricter power-law definition proposed by @huxley1932problems as there is ample evidence that this is appropriate for studies examining body proportions (CITE).

While both legacy rules arise from a similar theoretical concept, the analytical approaches used to analyze the two rules has differed substantially. Generally, Bergmann's rule is assessed using one or more regression models with a metric of body size (e.g., body mass) as the response variable in each model [e.g., @skinnerEnvironmentalGeographicConditions2025]. It is recommended to include multiple morphological metrics as complementary (though non-independent) representations of body size [@theriotHarnessingNaturalHistory2023]. While this may result in conflicting signals across traits [e.g., @baileyUsingDifferentBody2020], this interpretative complexity more accurately reflects the true nature of environment–morphology relationships.

Allen's rule, on the other hand, was formulated and has generally been interpreted as how appendage length changes *relative* to body size across environmental gradients [@allen1877influence, @rydingMultipleRegressionNot2025, @symondsItsAllRelative2025]. In other words, estimating relative appendage length requires tying appendage length and body size together *at the level of the individual* through some sort of allometric scaling relationship (@nte-allometry). This can be done in a multiple regression framework [@rydingMultipleRegressionNot2025], or by adhering to a power law definition [as recommended by @santoroAllometryEvaluateAllens2022, REEVALUATE THIS - WHAT EXACTLY DOES POWER LAW IMPLY?]. Thus, Allen's rule is inherently a multivariate endeavor while Bergmann's rule could theoretically be examined using a single metric of body size.

In other words, estimating relative appendage length requires tying appendage length and body size together *at the level of the individual* through some sort of allometric scaling relationship (@nte-allometry). This can be done in a multiple regression framework [@rydingMultipleRegressionNot2025], or by adhering to a power law definition [as recommended by @santoroAllometryEvaluateAllens2022, REEVALUATE THIS - WHAT EXACTLY DOES POWER LAW IMPLY?]. Thus, Allen's rule is inherently a multivariate endeavor while Bergmann's rule could theoretically be examined using a single metric of body size. The studies using metrics of *relative* appendage length, either implicitly or explicitly, acknowledge that allometric scaling is critical in the examination of Allen's rule [@santoroAllometryEvaluateAllens2022; @rydingMultipleRegressionNot2025; @symondsItsAllRelative2025].

OPTION: The majority of studies examining Bergmann's rule in isolation used multiple traits of body size. Examination of trait covariance (i.e., allometric relationships) was done in \~10% of studies. IF this is true, could then speculate why this difference arose. Bergmann's rule researchers are thinking about body size - be it approximated with appendage, mass, total length, whereas Allen's rule researchers are explicitly examining trends in body size across ecological gradients (in the context of examining relative appendage length).

Key here is that analyzing absolute appendage is not necessarily WRONG.

::: {#nte-allometry .callout-note}
## Allometric scaling in studies assessing relative appendage length

Allometry is the study of how biological processes change with body size. Seminal work by @huxley1932problems formulated that the scaling relationship between an appendage $Y$ and body size $X$ follows a power-law, where

$$
Y = \alpha * X^\beta
$$

On the log scale this function becomes additive

$$
log(Y) = log(\alpha) + \beta * log(X)\
$$

where $log(\alpha)$ determines the average relative appendage length for a given body size (the allometric intercept), and $\beta$ determines the rate that the appendage increases with an increase in body size [the allometric slope, @shingletonAllometryStudyBiological2010].

Allometric scaling theory posits that proportional growth of two linear features should increase linearly ($\beta$ = 1), whereas proportional growth of a linear feature and volume should scale inverse cubically [$\beta = 1/3$ i.e. given that organismal volume is three-dimensional, @raynerFormFunctionAvian1988]. When empirical reality matches theoretical expectations, this is known as isometric scaling [@shingletonAllometryStudyBiological2010]. If the allometric slope exceeds that expected by theory (e.g. $\beta > 1/3$), this is known as hyperallometry (or positive allometry), and when it is is less than expected by theory (e.g. $\beta < 1/3$), this is known as hypoallometry. Importantly, isometric scaling implies that body *shape* remains constant as body size increases, whereas hyper- and hypo- allometry suggest changes in body proportions.
:::

{{< pagebreak >}}

# References
