---
title: "Beyond Bergmann and Allen: A Unified Framework for Thermoregulatory Morphology"
# If you do not want code to show... echo: false
execute: 
 echo: false
 message: false
 warning: false
 cache: false
format: #docx
  elsevier-pdf: # elsevier-pdf
    header-includes:
      - |
        \usepackage{fancyhdr}
        \pagestyle{fancy}
        \fancyhead[R]{Thermo-Morpho Rule in a Warming World}
        \usepackage{makecell}
        \usepackage{lineno}
        \linenumbers
        \usepackage[margin=1in]{geometry}
        \setlength{\LTleft}{\fill}
        \setlength{\LTright}{\fill}
    journal:
      name: Global Ecology and Biogeography
      formatting: doubleblind
      cite-style: authoryear
editor: visual
bibliography: ../Suppfiles/Morphology.bib
metadata-files: 
  - ../Suppfiles/abstract_keywords_geb.yml
editor_options: 
  chunk_output_type: console
---

# Keywords

Bergmann's rule, Allen's rule, thermoregulation, thermal ecology, ecogeography, Thermo-morpho rule, climate change adaptation, complementarity, trade-offs, review

{{< pagebreak >}}

# Introduction

We are in the Anthropocene — a period of rapid global change dominated by habitat loss and climate change. While many species have not kept pace with these changes and have gone extinct, others have altered their shapes, sizes, distributions, and behaviors to adjust to this new human-dominated world [@parmesanEcologicalEvolutionaryResponses2006; @gardnerDecliningBodySize2011; @millienEcotypicVariationContext2006]. Understanding these sub-lethal responses is critical to understanding humanities’ full impact on biodiversity [@guoApplyingBergmannsRule2025]. With this knowledge, we will be better able to forecast changes to the natural world, as well as their impacts on ecosystem service provisioning [@luckImprovingApplicationVertebrate2012].

An organism’s morphology is perhaps the most fundamental determinant of how it interacts with its surroundings, and thus there are strong selective pressures to ensure that morphology is well-suited to the environment [@pigotMacroevolutionaryConvergenceConnects2020]. Thus, scientists have long sought to understand how morphology is shaped by environmental gradients [e.g., @winklerEcomorphologyMigrants1992], and some of the oldest macrogeographical 'rules' attempt to make sense of patterns in body size and shape. For example, Bergmann’s rule posits that individuals or populations of endothermic species tend to be larger in colder climates, presumably to minimize heat loss via a reduced surface-area-to-volume ratio [@bergmann1848ueber]. The related Allen’s rule states that appendages (e.g., limbs, tails, bills) tend to be shorter in colder regions, again reducing heat loss via a decreased surface-area-to-volume ratio [@allen1877influence].

There is general support for both Bergmann’s and Allen's rules in endotherms [@teplitskyClimateWarmingBergmanns2014; @rydingShapeshiftingChangingAnimal2021; @meiriValidityBergmannsRule2003], although there are also several important exceptions [@stevensonAllensRuleNorth1986; @geistBergmannsRuleInvalid1987; @cuiFunctionalAdaptationRather2020; @alhajeriGlobalTestAllens2020]. Bergmann's rule is perhaps the most extensively studied macroecological rule to date, and several important reviews cover adherence to [and lack thereof, @ashtonPatternsWithinSpeciesBody2002; @riemerNoGeneralRelationship2018], potential mechanisms of [@yom-tovRecentSpatialTemporal2011], and its implications in the face of climate change [@gardnerDecliningBodySize2011; @teplitskyClimateWarmingBergmanns2014]. Allen's rule is less well studied [but see @rydingShapeshiftingChangingAnimal2021], despite evidence that it is equally important for thermoregulation [@frohlichAllometryRevealsTradeoffs2023].

Importantly, both rules stem from a central tenet: changes in body size or shape alter an organism’s surface-area-to-volume ratio, with direct consequences for thermoregulation. Metabolic scaling theory shows that heat production generally scales with body volume to the 3/4 power, while the area available for heat exchange scales with volume to the 2/3 power (assuming isometric growth, i.e., no change in shape). Thus, as size increases, heat production outpaces heat dissipation, unless the organism alters its shape to become proportionally longer [@kleiber1932body; @kleiber1972body; @glazier2008effects]. The fact that size and shape both serve thermoregulatory functions and that they may act in concert to achieve thermoregulatory goals strongly suggests that they should be analyzed jointly (@fig-sa-to-v). Although the following discussion also extends to other environmental gradients that influence homeostasis through organismal surface-area-to-volume ratios [e.g., precipitation @jamesGeographicSizeVariation1970; @jirinecMorphologicalConsequencesClimate2021; @paltrinieriEffectsClimateBat2025], I will focus on thermoregulation for brevity.

![**Complementarity in Bergmann's and Allen's rules with increasing temperatures.** In response to increasing temperatures, organisms can achieve changes to their surface area to volume ratios by altering their a) body size, b) relative appendage size, c) or through complementarity, i.e. more minute changes to both body and appendage size. In c), morphological responses in body and appendage size are less pronounced than expected if Bergmann's and Allen's rules were operating in isolation. Figures were created by simulating changes to a cartoon bird made up of two spheres (representing the body and head) and a cone (representing the bill). Figure reproduced with permission from Baldwin et al (2023), see article for additional details.](../Figures/Baldwin_fig2.png){#fig-sa-to-v}

# Literature review

I queried the Web of Science to determine how often size (Bergmann's rule) and shape (Allen's rule) are examined a) jointly versus in isolation, and b) in an integrated fashion versus in parallel. I define an integrated analysis as examining some sort of interaction between size, shape, and temperature, or examining the relationships between temperature and size (i.e. $\beta_{size}$) and temperature and shape (i.e. $\beta_{shape}$). I considered an analysis as conducted in parallel if environment-morphology (size and shape) relationships were never considered in a single model. Part b) was conducted on the subset of studies examining size and shape jointly.

I limited my search to empirical studies published in the past 5 years (2021-2025) to capture current analytical practices and to assess whether joint analyses have become more common following the publication of influential work highlighting the importance of a joint analytical approach [@baldwinComplementarityAllensBergmanns2023; @frohlichAllometryRevealsTradeoffs2023]. To be included in the review, the study had to examine at least one continuous morphological variable in response to an environmental gradient (or a proxy, e.g., latitude, elevation, year) in an extant species or group of birds or mammals. Only studies focusing on Bergmann's or Allen's rules were considered, which I gauged by the inclusion of the word 'Bergmann\* or Allen\* at least once in the abstract, keywords, introduction, or methods.

The final search was conducted on May 8th, 2026, using the following search terms: (TS=(Allen\* OR Bergmann\* OR "ecogeographic\* rule*" OR "biogeographic* rule" OR "morpholog\* change*") AND TS=(*bird\* OR avian OR aves OR passerine\* OR primate\* OR rodent\* OR carnivore\* OR bat\* OR chiroptera OR "small mammal*" OR mammal* OR endotherm\* OR homeotherm*) AND TS=(temperature OR "thermal adaptation" OR latitud* OR elevation\* OR year\* OR thermoregulat\* OR "climat\* change" OR "global change" OR "climat\* warming" OR "environmental gradient" OR temporal*)) NOT TS=(fossil* OR paleontolog\* OR Cretaceous OR Jurassic OR Miocene OR Pleistocene) AND DT=(ARTICLE OR "EARLY ACCESS") AND WC=(ECOLOGY OR ZOOLOGY OR "EVOLUTIONARY BIOLOGY" OR BIOLOGY OR "VETERINARY SCIENCES" OR "BIODIVERSITY CONSERVATION" OR ORNITHOLOGY OR "ANATOMY MORPHOLOGY" OR PHYSIOLOGY).

```{r}
#| label: Setup

# Load libraries
library(tidyverse)
library(readxl)
library(janitor)
library(smatr)
library(ggpubr)
library(cowplot)
library(broom)

# Source functions
source("Scripts/Key_allometry_fns.R")
```

```{r}
#| label: Load-in-format-data

# Read in and format
Lr_allens <- read_excel("Lit_review/Methods_ecogeographic_rules_5.7.26.xlsx", 
                         sheet = "Allens") %>% 
  clean_names() %>% 
  filter(include_exclude != "Exclude") %>%
  dplyr::select(
    authors, year, journal, primary_question, n_spp, class, bergmann, interaction_size_shape, url_or_doi
    ) %>% 
  mutate(focus = ifelse(bergmann == "Y", "Both", "Allens_only")) %>% 
  dplyr::select(-bergmann)

# From Recs_allens.qmd script
Allens_mv <- read_csv("Lit_review/Multivariate.csv")
# Join and turn NAs into 'No'
Lr_allens2 <- Lr_allens %>% left_join(Allens_mv) %>% 
  mutate(multivariate = ifelse(is.na(multivariate), "No", multivariate))

Lr_bergs <- read_excel("Lit_review/Methods_ecogeographic_rules_5.7.26.xlsx", 
                         sheet = "Bergmanns") %>% 
  clean_names() %>% 
  filter(include_exclude != "Exclude") %>%
    dplyr::select(
    authors, year, journal, primary_question, class, allens, url_or_doi
    ) %>%
  mutate(focus = ifelse(allens == "Y", "Both", "Bergmanns_only")) %>% 
  dplyr::select(-allens)

# Bind together 
Lr_ecogeo <- bind_rows(Lr_allens2, Lr_bergs)
```

```{r}
#| label: extract-doi

# Function to extract DOI from a URL or string
extract_doi <- function(x) {
  str_to_lower(
    str_extract(x, "10\\.\\d{4,9}/[-._;()/:A-Za-z0-9]+")
  )
}

Lr_ecogeo2 <- Lr_ecogeo %>% mutate(doi_clean = extract_doi(url_or_doi))
```

```{r}
#| label: WOS-query
#| include: false

# Load in references from Web of science (WOS)
Wos_refs <- read_excel("Lit_review/Saved_recs_wos.xls") %>%
  clean_names() %>% 
  rename(journal = source_title,
         title = article_title, 
         year_publication = publication_year) %>%
  dplyr::select(authors, title, journal, year_publication, doi, author_keywords) %>%
  mutate(doi_clean = extract_doi(doi))

# Examine which references were not picked up by the WOS search
missing_refs <- Lr_ecogeo2 %>%
  anti_join(Wos_refs[,c("doi_clean")], #, "author_keywords", "keywords_plus"
             by = "doi_clean") #%>%
  #mutate(found_in_WoS = !is.na(article_title))

## Missing
missing_refs 
nrow(missing_refs) # Missing refs
nrow(missing_refs) / nrow(Lr_ecogeo2) # Proportion missing 
```

```{r}
#| label: Subset-lr-dfs

Lr_wos <- Lr_ecogeo2 %>% 
  inner_join(Wos_refs[,c("doi_clean")]) %>% 
  mutate(wos_search = "Yes")
Lr_homeostasis <- Lr_ecogeo2 %>% filter(primary_question == "Homeostasis")
```

```{r}
#| label: Calc-props-fn
#| include: false

# Create table for labeling with proportion in each category uh
Calc_props <- function(df, by_var) {
  df %>% 
  summarize(num_studies = n(), .by = c({{ by_var }})) %>% 
  mutate(Tot_studies =  sum(num_studies), 
         Prop = num_studies / Tot_studies,
         Per = Prop * 100) %>% 
  mutate(Per = round(Per, 0),
         Prop = round(Prop, 2), 
         Per = paste0(Per, "%"),
         {{ by_var }} := str_replace({{ by_var }}, "_", "\n"))
}
```

```{r}
#| label: Prop-tbls-full-lr
#| include: false

Prop_all <- Calc_props(Lr_ecogeo, by_var = focus)
Prop_wos <- Calc_props(Lr_wos, by_var = focus)

# NOTE the ratios do change if you include only studies that examine organismal homeostasis 
Calc_props(Lr_homeostasis, by_var = focus)
```

```{r}
#| label: interactions-dfs
#| include: false

# Of species that examined both Bergmann's and Allen's rules.. 
Lr_both <- Lr_ecogeo %>% 
  filter(focus == "Both") 

# Of studies that had at least 20 species.. This is important as the likelihood of examining Bergmann's and Allen's rules jointly increases greatly in multi-species studies
N_spp_cutoff <- 20
Lr_multi_spp <- Lr_both %>% 
  filter(n_spp >= N_spp_cutoff)

# Use Calc_props() 
Prop_both <- Calc_props(Lr_both, by_var = interaction_size_shape)
Prop_both_ms <- Calc_props(Lr_multi_spp, by_var = interaction_size_shape)
N_studies_multi_spp <- nrow(Lr_multi_spp)
```

```{r}
#| label: hist-num-spp
#| include: false

# Number of species in studies
Lr_both %>% filter(n_spp < 2000) %>%
  mutate(interaction_size_shape = case_when(
    interaction_size_shape == "N" ~ "Parallel", 
    interaction_size_shape == "Y" ~ "Integrated"
  )) %>%
  ggplot() +
  geom_histogram(color = "black", aes(x = n_spp, fill = interaction_size_shape)) + 
  scale_x_log10() + 
  labs(x = "Number of species", y = "Count", fill = " ") + 
  scale_fill_grey()
ggsave("Figures/Integrated_parallel_hist.png", bg = "white")
```

```{r}
#| label: pull-per-fns

pull_per <- function(df, filter_by, focus){
  df %>% filter({{ filter_by }} == {{ focus }}) %>% 
    pull(Per)
}
calc_per_iso <- function(df){
  df %>% filter(focus != "Both") %>% 
    summarize(
      Isolated_prop = scales::percent(
        sum(num_studies) / unique(Tot_studies), accuracy = 1
        )
      ) %>% pull(Isolated_prop)
}
```

```{r}
#| label: pull-percents-full-lr

# Use custom functions to extract percentages 
Per_allen <- pull_per(df = Prop_all, filter_by = focus, focus = "Allens\nonly")
Per_berg <- pull_per(df = Prop_all, filter_by = focus, focus = "Bergmanns\nonly")
Per_both <- pull_per(df = Prop_all, filter_by = focus, focus = "Both")
Per_iso <- calc_per_iso(Prop_all)
Per_iso_wos <- calc_per_iso(Prop_wos)
 
# Percent of studies that examined Allen's as relative appendage length
Prop_ral <- nrow(Allens_mv) / nrow(Lr_allens2)
Per_num <- (Prop_ral * 100) %>% round(0)
Per_ral <- paste0(Per_num, "%")
```

```{r}
#| label: pull-per-interactions-dfs
#| include: false

Per_interact <- pull_per(Prop_both, filter_by = interaction_size_shape, focus = "Y")
Per_no_interact <- pull_per(Prop_both, filter_by = interaction_size_shape, focus = "N")

# Multispecies studies (> 10 species)
Per_interact_ms <- pull_per(Prop_both_ms, filter_by = interaction_size_shape, focus = "Y")
```

```{r}
#| label: pull-percents-class

pull_per_class <- function(class){
  Per_class %>% filter(class == {{ class }}) %>% 
  pull(percent)
}

# Percent of each class (aves, mammalia, or both)
Per_class <- Lr_ecogeo %>% tabyl(class) %>% 
  mutate(percent = round(percent, 2) * 100, 
         percent = paste0(percent, "%")) 

Per_birds <- pull_per_class("Aves")
Per_mammals <- pull_per_class("Mammalia")
Per_both <- pull_per_class("Aves and mammalia")
```

```{r}
#| label: wos-review-nums

Num_elim <- nrow(Wos_refs) - nrow(Lr_wos)
Num_additional_refs <- nrow(Lr_ecogeo2) - nrow(Lr_wos)
```

The initial query produced `r nrow(Wos_refs)` studies which were manually reviewed to ensure conformity to the inclusion criteria, resulting in the elimination of `r Num_elim` studies. During processing of the remaining `r nrow(Lr_wos)` articles, I found `r Num_additional_refs` additional studies that met the criteria for inclusion but had not been identified by the original query, yielding a final corpus of `r nrow(Lr_ecogeo2)` studies used in the current review. The majority of these studies were conducted in birds (`r Per_birds`), while mammals were examined in `r Per_mammals` of studies, and `r Per_both` examined both taxa (Supporting information).

Of `r nrow(Lr_ecogeo2)` studies examining Bergmann's or Allen's rules, `r Per_iso` examined size and shape in isolation (@fig-barplots). Results were identical when restricting the analyses to just those identified in the Web of Science search; hereafter, I only report results from the larger set of `r nrow(Lr_ecogeo2)` studies. Importantly, authors continue to analyze ecogeographic rules in isolation even after the publication of two seminal papers in 2023 [@frohlichAllometryRevealsTradeoffs2023; @baldwinComplementarityAllensBergmanns2023]. A focus on Bergmann's rule alone (`r Per_berg`) was far more common than on Allen's rule alone (`r Per_allen`, @fig-barplots), likely because Allen's rule studies generally measured *relative* appendage length (`r Per_ral` of studies; Supporting information), i.e. appendage length controlling for body size [@symondsItsAllRelative2025]. Thus, in some respect, studies of Allen's rule were already considering body size across environmental gradients.

Importantly, of the `r nrow(Lr_both)` studies that analyzed both Allen's and Bergmann's rules, only `r Per_interact` examined the rules in an integrated fashion; the other `r Per_no_interact` examined the two rules in parallel. The number of focal species in the study influenced how often authors conducted integrated versus in parallel analyses. All studies with under 20 focal species conducted analyses in parallel, whereas half of the studies with at least 20 focal species conducted integrated analyses (@fig-barplots). It is important to note some types of integrated analyses are achievable with any number of focal species (e.g., interaction models such as in @fig-interactions), while others are not [e.g., metaregression, as in @paltrinieriEffectsClimateBat2025; @baldwinComplementarityAllensBergmanns2023].

```{r}
#| label: export-csv 
#| include: false

# Format and export csv with codings of studies
Lr_export <- Lr_ecogeo2 %>% 
  anti_join(Lr_wos) %>% 
  mutate(wos_search = "No") %>%
  bind_rows(Lr_wos) %>% 
  dplyr::select(-url_or_doi) %>% 
  rename(doi = doi_clean, 
         relative_appendage = multivariate) %>% 
  arrange(desc(wos_search), authors) %>% 
  mutate(
    primary_question = str_remove_all(primary_question, "Homeostasis;?\\s*")
    ) %>%
  # For now, remove primary_question column. I don't think I was standardized enough in coding the studies. A better methodology might be to just code a binary column whether thermoregulation (or homeostasis?) was one of the primary goals of the study. If the study had terms X Y or Z it counted as one of the primary questions.
  dplyr::select(-primary_question) 

# Maintain accents using readr::write_excel_csv
#Lr_export %>% write_excel_csv("Journal_submissions/Lit_review_berg_allen.csv")

## Export refernces from Web of Science 
if(FALSE){
  Wos_refs %>% 
  dplyr::select(-doi) %>%
  mutate(Comments = ifelse(
    str_detect(authors, "Gyurácz"), "No DOI, Article No.: e231602", NA)
    ) %>% write_excel_csv("Journal_submissions/Wos_literature_search.csv", na = " ")
}
```

```{r}
#| label: barplots-figure
#| include: false

# Format
Lr_glm <- Lr_ecogeo2 %>% 
  mutate(focus_label = str_remove(focus, "_only"),
         focus_label = ifelse(focus_label == "Both", "Joint", focus_label),
         focus_bin = ifelse(focus == "Both", 1, 0),
         before_after = ifelse(year > 2023, "After", "Before"))

# Generate the proportion of each focus type ('Both', 'Allens', 'Bergs')
Lr_prop_before_after <- Lr_glm %>% 
  count(before_after, focus_label) %>% 
  mutate(Per = n / sum(n),
         panel = "b",
         .by = before_after)
#Lr_prop_before_after %>% summarize(sum(n), .by = before_after)

# Generate proportions
Lr_prop_all <- Lr_glm %>% 
  count(focus_label) %>% 
  mutate(Per = n / sum(n),
         before_after = "All studies",
         panel = "a")

# Bind together for single df for plotting
Lr_props <- bind_rows(Lr_prop_all, Lr_prop_before_after) %>% 
  mutate(before_after = factor(
    before_after, levels = c("All studies","Before","After")
    ))

# Model and extract p-value
mod <- glm(focus_bin ~ before_after, data = Lr_glm, family = "binomial")
p_val <- broom::tidy(mod) %>% 
  filter(term == "before_afterBefore") %>% 
  pull(p.value) %>% 
  round(2)

## Plot
# Generate tbl with information to create a) and b) labels for geom_text()
labels_df <- tibble(
  panel = c("a", "b", "b"),
  x = c(1 - 0.5, 1 - 0.54, 1 + .3),
  y = c(1.08, 1.08, 1.08),
  lab = c("a)", "b)", paste("p =", p_val)),
  size = c(6, 6, 4)   # smaller font for p-value
)

# Plot
Lr_props %>% 
  ggplot(aes(x = before_after, y = Per, fill = focus_label)) +
  geom_col(color = "black") +
  geom_text(aes(label = round(Per, 2)),
            position = position_stack(vjust = 0.5), size = 4) +
  facet_wrap(~panel, scales = "free_x") +
    geom_text(data = labels_df,
            aes(x = x, y = y, label = lab, size = size),
            inherit.aes = FALSE, hjust = 0) +
  scale_size_identity() +
  scale_fill_manual(
    values = c("Allens"="white","Bergmanns"="gray70","Joint"="gray30")
    ) +
  labs(x=NULL, y="Proportion", fill=NULL) +
    theme(
    strip.background = element_blank(),
    strip.text = element_blank(),
    legend.position = "top"
  )
#ggsave("Figures/Allen_berg_both_barplot.png", bg = "white")
```

![The frequency with which Allen’s and Bergmann’s rules were analyzed in isolation vs jointly. a) Across all studies (n = 87), the two rules were analyzed jointly less than half of the time (47%). The inset histogram shows that, of the joint analyses, the majority of them tested the two rules in parallel instead of in an integrated fashion, although the number of focal species moderated this relationship. Not pictured are the Fröhlich et al and Baldwin et al. studies, which both conducted integrated analyses and included 9,962 and 6,974 species, respectively. b) Logistic regression showed no increase in the frequency with which Allen’s and Bergmann’s rules were anlyzed jointly before (n = 49) vs after 2023 (n = 38). The end of 2023 was chosen due to the publication lag after seminal papers by Fröhlich et al and Baldwin et al. were published in February and July of 2023, respectively. Thus, 'Before' includes the studies published in 2023.](../Figures/Barplot/Barplot_allen_berg_both.pdf){#fig-barplots}

# The need for a combined approach

Fewer than half of the studies published in the last five years examined Bergmann’s and Allen’s rules together, and even among these, the vast majority tested them in parallel rather than in an integrated framework. It is therefore important to consider what is lost when size and shape are studied as separate phenomenon and the possibilities of dynamic interplay are ignored. I argue that size, shape and temperature gradients regularly interact with each other, and thus require an integrated analytical framework to make robust inference. I discuss interactions broadly and then focus on two types of correlated responses that are highlighted in recent global analyses: trade-offs and complementarity [@baldwinComplementarityAllensBergmanns2023; @frohlichAllometryRevealsTradeoffs2023].

## Interactions: size, shape, and temperature

The joint examination of Bergmann's and Allen's rules allows for a more nuanced interpretation of morphology-environment relationships, particularly by testing interactions between size, shape, and temperature. For example, recent work from the new and old world and across both hemispheres shows that smaller species tend to exhibit stronger morphological responses to warming than larger species [@mcqueenShorebirdsAreShrinking2024; @neate-cleggAfromontaneUnderstoryBirds2024; @paltrinieriEffectsClimateBat2025; @zimovaBodySizePredicts2023]. This could be because smaller species are exposed to more extreme thermoregulatory pressures, are less buffered relative to larger species, or because they exhibit greater evolutionary potential due to larger effective population sizes, higher genetic variance, and faster molecular evolution [@orkneySmallBodySize2024; @zimovaBodySizePredicts2023].

Importantly, the majority of the aforementioned studies used a two-step approach to understand how body size influenced the rate of morphological change (e.g., metaregression analysis). This approach is tremendously valuable, but does not directly allow for moderation of the estimand of interest (e.g., $\beta_{temperature}$ or $\beta_{year}$) based on combinations of size and shape. @frohlichAllometryRevealsTradeoffs2023 illustrated the power of this approach by including interactions between size, shape, and temperature in a single model. They found that temperature–morphology relationships were dependent on bird size and shape, and that certain morpholgical combinations reversed overall patterns conforming to Bergmann's and Allen's rules. For example, despite overall conformity to Bergmann’s rule, inverse patterns (i.e., smaller body size at colder temperatures) emerged in birds with particularly short bills and long legs. Likewise, Allen’s rule only held for large-bodied species whereas small-bodied species showed the opposite trend (@fig-interactions).

![Modeling interactions between size and shape is critical in understanding tradeoffs in organismal response to increasing temperatures. In predicting a species' experienced temperature envelope, Fröhlich et al (2023) found that the environment-morphology relationship of mass (top left), beak (top right), and tarsus (bottom) was dependent on the size and shape of the organism. All relationships were highly significant at p \< 0.001. See Figure 4 in Fröhlich et al (2023) for additional infromation; figure was modified and reproduced with permission from Arkadiusz Fröhlich.](../Figures/Frohlich_fig4F_modified.pdf){#fig-interactions}

The general conceptual model is depicted in @fig-conceptual-model (panel a), where a species' size or shape influences how species respond morphologically across temperature gradients. Importantly, in all the studies examined in the present review, morphology-temperature interactions were tested across species and not within species. The brand of morphology-temperature interactions tested by @frohlichAllometryRevealsTradeoffs2023 could be examined within-species, and I believe this is an important area of future research.

![Temperature's influence on body size and shape is moderated by a) size and shape themselves, and b) through functional constraints. Importantly, in b) the effects of temperature on size are correlated with the effects of temperature on shape, as suggested by Baldwin et al (2023). Solid lines represent direct effects and straight dashed lines represent interactions (moderating the effect that they point to). The double headed curved arrow in b) represents correlated effects. Size influences shape through allometry.](../Figures/Conceptual_model/Conceptual_model_panels.pdf){#fig-conceptual-model}

## Trade-offs and complementarity: correlated responses in size and shape

Recent work has shown that morphological responses to temperature are often correlated (@fig-conceptual-model, panel b), such that stronger adherence to Bergmann's rule is often linked with weaker adherence to Allen's rule, and vice versa. Two recent integrated analyses of Bergmann's and Allen's rules have noted two different manifestations of this phenomenon. @frohlichAllometryRevealsTradeoffs2023 found support for the ‘trade-off hypothesis’, that species responded to warming either by decreasing body size or by increasing appendage length, but rarely both [also see @dubinerWidespreadRecentChanges2022]. They framed this as an evolutionary compromise, where the ecological cost associated with a certain thermoregulatory strategy (i.e., change in size or change in shape) is more or less viable depending on a species’ life history. @baldwinComplementarityAllensBergmanns2023 found evidence for 'complementarity', where coordinated shifts in both body size and appendage length jointly altered surface area–to–volume ratios to a similar extent as larger shifts in either trait alone (@fig-sa-to-v). Indeed, 81% of the 107 bird families examined did not conform with Bergmann’s or Allen’s rules when analyzed separately; instead, most families exhibited modest but coordinated changes in size and shape in the directions predicted by both rules [also see @dubinerWidespreadRecentChanges2022].

While the authors interpret their results differently, they can be reconciled within a common conceptual framework. Both analyses conclude that temperature-morphology (size and shape) responses are not independent of each other, and that changes in size and shape can incur fitness costs. For example, smaller body sizes can reduce competitive ability [@shipleySelectionCounteractsDevelopmental2022] or foraging efficiency [@vangilsBodyShrinkageDue2016], and changes to appendage size or shape can alter vocal signaling [@friedmanEvolutionMultifunctionalTrait2019; @sebastianelliContinentwidePatternsSong2022]. Both studies argue that these fitness costs can constrain morphological responses to temperature, such that life-history requirements moderate the relative extent of size versus shape change (@fig-conceptual-model, panel b). For example, bird bills appear to be functionally constrained in taxa with unique foraging ecologies (e.g., Trochilidae and Rallidae); thus, these taxa relied heavily on changes in body size to meet thermoregulatory needs [@baldwinComplementarityAllensBergmanns2023]. These studies therefore suggest that strong trade-offs may be more likely in organisms with highly specialized morphologies or strong functional constraints, whereas complementarity may dominate otherwise. Indeed, the coordinated but smaller shifts implied by complementarity may buffer against the fitness costs of extreme changes in a single trait while still permitting meaningful shifts to thermoregulatory capacity [@baldwinComplementarityAllensBergmanns2023]. Finally, both studies also noted that strong simultaneous adherence to Bergmann's and Allen's rules, in their classical forms, was rare [@frohlichAllometryRevealsTradeoffs2023] or non-existent [@baldwinComplementarityAllensBergmanns2023].

The similarities between these two theories highlight that they are not entirely distinct processes, but are better interpreted as correlated responses among functionally integrated traits. Let $\beta_{Tsize}$ and $\beta_{Tshape}$ represent the responses of body size and shape to temperature, respectively. Under this framework, trade-offs and complementarity differ primarily in how morphological response is partitioned across these two trait axes. In trade-offs, strong constraints on one trait axis may redistribute most thermoregulatory response onto the alternative axis, resulting in large changes in either size or shape, but rarely both simultaneously. In complementarity, constraints or fitness costs may be more evenly distributed across both axes, resulting in coordinated but smaller changes in both size and shape. Thus, rather than representing categorically different phenomena, trade-offs may instead represent a boundary case of complementarity, where response is highly concentrated along a single morphological dimension (@fig-tradeoffs-v-complementarity).

Importantly, both @baldwinComplementarityAllensBergmanns2023 and @frohlichAllometryRevealsTradeoffs2023 primarily attribute these integrated responses to functional constraints associated with species’ life histories. While this is likely an important mechanism, other processes could plausibly generate similar empirical covariance structures, including developmental integration, genetic covariance among traits, or phylogenetic inertia. Distinguishing among these alternative mechanisms remains important for interpreting whether observed covariance between $\beta_{Tsize}$​ and $\beta_{Tshape}$​ reflects adaptive compensation, correlated developmental responses, or shared evolutionary history. From a statistical perspective, these phenomena are most appropriately treated as correlated multivariate responses to temperature, i.e., non-independence between $\beta_{Tsize}$​ and $\beta_{Tshape}$, but conceptually, they can be thought of as a temperature × 'strength of constraint' interaction.

![Trade-offs and complementarty are distinct presentations of the same phenomena. In complementarity, species respond to changes in temperature via relatively small but coordinated changes in both size and shape. In trade-offs, the response to temperature is highly constrained in one functional trait (size in this case), so all morphological thermal response occurs through the other functional trait (shape in this case). The size and shape lines are interchangeable, such that the species could respond (more strongly) via changes in size instead of shape.](../Figures/Tradeoffs_v_complementarity.pdf){#fig-tradeoffs-v-complementarity}

## The Thermo-Morpho Rule

These examples show that trade-offs and complementarity are variations on a single theme – temperature's impact on morphology is nearly always moderated by other biological constraints. The strength of these constraints on size and shape determine the 'wiggle room' available for adaptation within each trait. Recognizing this shared structure highlights that interactions are the key to understanding morphological trends along temperature gradients. Thus, analyzing either rule in isolation precludes the possibility of testing for relationships *between* size and shape, despite that these codependencies are likely the norm in nature and not the exception to the rule. The overly simplistic thinking implicit in testing for either rule in isolation risks missing or even mischaracterizing the processes that underpin thermoregulation in most species.

The continued isolated treatment of Bergmann's and Allen's rules highlights the need for a framework shift. Here, I propose the Thermo-Morpho Rule, which views size and shape as complementary axes of thermoregulatory adaptation, providing theoretical justification for analyzing Bergmann’s and Allen’s rules in tandem. Under this framework, adherence occurs whenever shape and size interact in a way that increases surface area to volume ratios in warmer climates. This could be by following either Bergmann’s or Allen's rule (i.e., trade-offs), or exhibiting coordinated changes in body size and appendage length that together alter the surface area–to–volume ratio (i.e., complementarity).

# Beyond morphology: Toward an integrative thermoregulatory framework

Thermoregulation extends beyond morphology, as organisms respond to hotter climates through shifts in geographic distributions, physiology, behavior, or phenology. For instance, many species move upslope or poleward, advance breeding or migration timing, shift activity to cooler periods of the day, or adjust metabolic heat production [@millienEcotypicVariationContext2006; @mayrGeographicalCharacterGradients1956; @seebacherResponsesTemperatureVariation2009]. Ecogeographical rules beyond Bergmann’s and Allen’s also highlight alternative thermoregulatory pathways [@tianPredictingBioticResponses2020]. For example, Bogert’s rule (the “thermal melanism hypothesis”) proposes that darker pigmentation improves heat absorption in cool environments [@bogertThermoregulationReptilesFactor1949; @clusella-trullasTestingThermalMelanism2008], while Hesse's rule predicts that relative organ size will decrease if metabolic rates decline with warming temperatures [@mullerRelativeHeartSize2014; @seebacherResponsesTemperatureVariation2009]. Importantly, there has been work combining several of these rules [@macedoMacroevolutionBodyExtremities2025; @cardiliniTestsEcogeographicalRelationships2016; @cuiFunctionalAdaptationRather2020], but they have not incorporated the interactions between rules [but see @malpicaColorMorphologicalDifferentiation2022].

Similar to morphological rules (e.g., Bergmann's and Allen's), these alternative thermoregulatory pathways are likely dominated by interactions as well, such that one adaptive trait modifies how another trait responds to environmental temperature. For instance, in shorebirds, morphology influences behavioral thermoregulation: individuals with longer bills and legs more frequently exposed these appendages to facilitate dry heat transfer, while shorter-limbed species relied less on such behaviors [@pavlovicMorphologyGeographyPredict2019; @ryelandBillSizeMediates2017; @ryelandLegLengthTemperature2019; @playa-montmanyBehaviouralPhysiologicalResponses2023]. Similarly, migratory behavior interacts with temperature to shape body size clines – migratory species often show inconsistent responses to temperature, whereas resident species more reliably conform to Bergmann’s and Allen’s rules [@mainwaringConformityBergmannsRule2021; @mcqueenShorebirdsAreShrinking2024]. Other behaviors like torpor, (micro-) habitat selection, and nocturnality are important to consider as they can also influence the thermal environment experienced by wildlife [@brivioSeekingTemporalRefugia2024; @jirinecLightThermalNiches2022; @bonebrakeClimateChangeThermoregulatory2020]. Indeed, behavioural flexibility dampens the need for phenotypic responses [@baldwinPhenotypicResponsesClimate2022].

As with the morphological rules, alternative thermoregulatory pathways are subject to trade-offs and complementarity as well. In this context, trade-offs are when a given thermoregulatory pathway is highly constrained, leading to reliance on alternative mechanisms of thermoregulation. For example, koalas (*Phascolarctos cinereus*) exhibit limited evolutionary scope for body-size change because arboreal foraging favors small, lightweight bodies, but they compensate through geographic variation in pelage insulation that closely tracks temperature [@briscoeBergmannMeetsScholander2015]. Similarly, woodrats (Neotoma) lack key physiological adaptations for heat dissipation (e.g., efficient evaporative cooling), but show strong temperature–body size relationships suggesting that morphological adjustment substitutes for physiological adaptation [@brownBergmannsRuleClimatic1969; @smithPredictingWoodratNeotoma2006; @khaliqGlobalVariationThermal2014]. Complementarity occurs when multiple traits act together to maintain thermal balance. For example, @he2023global showed that mammals and birds that hibernate or migrate to avoid the coldest temperatures still adhered to Bergmann's rule, but less strongly than their counterparts who felt the full brunt of winter. Similarly, @mollerMorphologicalConstraintsChanging2017 showed that birds with high aspect ratios adjusted migration phenology less than those with lower aspect ratios. Higher aspect ratios may increase surface area relative to body mass, facilitating convective heat exchange [@rubalcabaPhysicalConstraintsThermoregulation2022] and reducing selective pressure for adjustments to migration timing.

These studies suggest that any trait under thermoregulatory selection could modify, act in concert with, or substitute for morphological adaptation [@clarkeScalingBasalMetabolic2010; @tianPredictingBioticResponses2020; @khaliqGlobalVariationThermal2014]. A natural next step is to formalize a broader *Thermo-Adaptive Framework*, encompassing how thermal gradients jointly shape morphology, physiology, coloration, and behavior [@benhamIntegratingSpatialAnalyses]. The Thermo-Adaptive Framework is best viewed as a conceptual scaffold linking thermoregulatory function across traits. I do not frame it as an explicit rule with predictions to test (although this is feasible); instead its purpose is to raise awareness of trade-offs, complementarity, and interactions amongst adaptive thermoregulatory traits. Such a framework will require data that are not yet routinely collected, but, as with morphological databases [@tobiasAVONETMorphologicalEcological2022], comparable datasets on physiological, behavioural, and color traits are rapidly emerging (e.g., [Project Plumage](https://www.nhm.ac.uk/take-part/monitor-and-encourage-nature/project-archive/project-plumage.html)). By integrating across these axes, we can move toward a more complete understanding of how organisms meet the thermoregulatory challenges of a warming world [@guoApplyingBergmannsRule2025].

# Taxonomic scope

With few exceptions [e.g., @briscoeBergmannMeetsScholander2015; @playa-montmanyBehaviouralPhysiologicalResponses2023], the work discussed here on interactions, trade-offs, and complementarity has been conducted within an interspecific framework. Yet intraspecific variation in morphology can be substantial (CITE Stevens 2010, seagull study), raising the question of why interactions among temperature, size, and shape have rarely been examined within species. Although some intraspecific questions remain difficult to address, data availability alone is unlikely to explain the relative scarcity of within-species studies. For example, the interaction framework examined by @frohlichAllometryRevealsTradeoffs2023 could just as easily be tested within species given adequate sample sizes. Furthermore, advances in biologging and movement ecology have dramatically increased the availability of individual-level data on movement, physiology, behavior, and morphology [CITE KAYS, 2015; Wilmer, 2015 golden age of bio-logging; Jetz TREE paper, 2022; Motus tracking]. We have arguably been in a “golden age of biologging” for more than a decade, yet these data streams remain underutilized to examine intraspecific interactions despite their great potential. For example, biologging data can transform coarse species-level categories such as “migratory” versus “non-migratory” into continuous individual-level measures such as migration distance, which can then be paired with morphology to test how migratory behavior influences adherence to ecogeographical rules [e.g., @skinnerEnvironmentalGeographicConditions2025]. Intraspecific analyses may also help disentangle mechanisms that are difficult to separate in comparative datasets, including developmental integration, phenotypic plasticity, behavioral buffering, and evolutionary constraint. I hope the recommendations provided here prove useful for future intraspecific analyses, which may be particularly valuable for identifying populations whose morphological responses are constrained or mismatched under rapid climatic change.

# Conclusions

Substantial resources go into collecting and analyzing phenotypic data to test ecogeographical rules. Yet more than half of the studies published in the last five years examined Bergmann’s and Allen’s rules in isolation, limiting our understanding of how size and shape co-vary across environments. In the Anthropocene, where conservationists must make fast and informed decisions, the continued compartmentalization of trait variation into separate “rules” is increasingly unhelpful [@goldenbergLinkBodySize2022]. Analyzing multiple traits simultaneously within an integrated framework provides greater inferential power and a more realistic picture of how phenotypes respond to temperature and other environmental gradients. Indeed, interactions are important in understanding thermoregulation [@lustickInteractionPostureColor1980] and ecogeographical rules more broadly [@gastonEcogeographicalRulesElements2008; @marcondesRethinkingGlogersRule2021; @delheyReconcilingEcogeographicalRules2019]. These insights matter at both intra- and inter- specific scales: intraspecific analyses can forecast how species will respond to rising temperatures within their ranges and identify vulnerable populations, whereas interspecific analyses can reveal how entire clades respond and highlight groups of organisms particularly susceptible to warming. Ultimately, identifying the pathways - whether morphological, physiological, behavioral, or chromatic - through which selection acts will enable conservationists to better anticipate biological responses in an increasingly warm and human-dominated world [@benhamIntegratingSpatialAnalyses].

# Data availability

The full set of `r nrow(Wos_refs)` articles from the Web of Science search, as well as my coding of the `r nrow(Lr_ecogeo2)` studies that met the inclusion criteria, are available for download in the supporting information.

# EXTRAS

Save for a few specific examples \[@briscoeBergmannMeetsScholander2015; @playa-montmanyBehaviouralPhysiologicalResponses2023\], the work presented here examining interactions, trade-offs, and complementarity was conducted in an interspecific framework. The variation within species is often non-negligible (CITE Stevens 2010, seagull study), so this begs the question 'why have intraspecific interactions been largely ignored?' While certain intraspecific questions may be beyond the data that is readily available, this is not generally the limiting factor. For example, the brand of morphology-temperature interactions tested by @frohlichAllometryRevealsTradeoffs2023 could have been examined just as easily within-species provided adequate sample sizes. Furthermore, we have been in the 'golden age of biologging' (CITE KAYS, 2015; Wilmer, 2015 golden age of bio-logging, Jetz TREE paper, 2022; Motus tracking) for more than a decade, so individual-level data on movement, physiology, behavior, and morphology exists. Whatever the reason, this as an important area of future research. For example, GPS tracking can turn species-level traits like migration status (e.g., 'migratory' or 'non-migratory') into continuous individual-level variables like migration distance, and ultimately paired with morphology to understand how the degree of migration influences adherence to ecogeographical rules [e.g., @skinnerEnvironmentalGeographicConditions2025]. I hope the recommendations provided here also serve as a guide for scientists conducting intraspecific analyses, which may be important in identifying vulnerable populations as temperatures continue to rise and climates become more variable.

{{< pagebreak >}}

# References
