---
title: "Beyond Bergmann and Allen: A Unified Framework for Thermoregulatory Morphology"
number-sections: true
# If you do not want code to show... echo: false
execute: 
 echo: false
 message: false
 warning: false 
format: docx
editor: visual
bibliography: ../Supp_files/Morphology.bib
link-citations: true
csl: ../Supp_files/methods-in-ecology-and-evolution.csl
knitr:
  opts_chunk: 
    comment: "#>"
    collapse: true
editor_options: 
  chunk_output_type: console
---

{{< pagebreak >}}

# Abstract

**Background:** Understanding how organismal size and shape vary across environmental gradients is essential in ecology, evolution, and conservation science. Yet size and shape are often studied separately, despite their shared functional basis. Of 48 studies published since 2021 examining Bergmann's rule, Allen's rule, or both, nearly half examined Bergmann's or Allen's rule in isolation. A focus on Bergmann's rule in isolation was much more common than focusing on Allen's rule in isolation.

**A unified framework:** I propose replacing the siloed treatment of Bergmann’s and Allen’s rules with a single conceptual model: the Thermo-Morpho Rule. The Thermo-Morpho Rule views size and shape as complementary axes of thermoregulatory adaptation, both governed by the same biophysical principle – that surface area to volume ratios mediate heat exchange. Two recent joint analyses of Bergmann’s and Allen’s rules support this framework, highlighting the complexity of morphological response to environmental change. Indeed, species responded to warming 1) either by decreasing in body size or by increasing relative appendage length, but rarely both; and 2) groups exhibited small, coordinated shifts in both traits that together achieved meaningful change to the surface area to volume ratio. \[MAYBE IT DOESN'T MAKE SENSE TO NAME THIS RULE. IF THE *Thermo-Adaptive Rule* IS MORE WHERE I WANT TO TAKE THE PAPER?\]

**Future work:** The adaptive landscape of temperature extends beyond morphology. Other ecogeographical rules related to thermoregulation (e.g., Bogert's or Hesse's rules) along with behavioral and physiological adjustments, may act in complementary ways to morphology. These patterns suggest the need for a broader ‘Thermo-Adaptive Rule,’ which integrates multiple axes of phenotypic response to temperature gradients. I highlight recent studies that use integrated models to examine multiple ecogeographical rules simultaneously and argue that this should be the dominant theoretical and analytical paradigm moving forward.

**Conclusions:** Understanding phenotypic responses to climate change requires moving beyond legacy rules and toward integrative models that examine a broader range of thermoregulatory adaptation. In an era of rapid global change, conservationists can’t afford to split hairs over legacy rules and instead should adopt frameworks that recognize the interdependence of traits, function, and environment.

# Introduction

We are in the Anthropocene — a period of rapid global change dominated by habitat loss and climate change. While many species have not kept pace with these changes and have gone extinct, other species have been able to alter their shapes, sizes, ranges, and behaviors to adjust to this new human-dominated world [@parmesanEcologicalEvolutionaryResponses2006; @gardnerDecliningBodySize2011; @rydingShapeshiftingChangingAnimal2021]. Understanding the sub-lethal, plastic or evolutionary responses is critical to understanding humanities’ full impact on biodiversity [@guoApplyingBergmannsRule2025]. With this knowledge, we will be better able to forecast changes to the natural world, as well as their impacts on ecosystem service provisioning.

An organism’s morphology is perhaps the most fundamental determinant of how it interacts with its environment, and thus there are strong selective pressures to ensure that morphology is well-suited to its environment [@pigotMacroevolutionaryConvergenceConnects2020]. Thus, scientists have long sought to understand how morphology is shaped by the environment [e.g., @winklerEcomorphologyMigrants1992], and some of the oldest macrogeographical 'rules' attempt to make sense of patterns in body size and shape. For example, Bergmann’s rule posits that individuals or populations of endothermic species tend to be larger in colder climates, presumably to minimize heat loss via a reduced surface-area-to-volume ratio [@bergmann1848ueber]. The complementary Allen’s rule states that appendages (e.g., limbs, tails, bills) tend to be shorter in colder regions, again reducing heat loss via a decreased surface-area-to-volume ratio [@allen1877influence].

There is general support for both Bergmann’s and Allen's rules in endotherms [@teplitskyClimateWarmingBergmanns2014; @rydingShapeshiftingChangingAnimal2021; @meiriValidityBergmannsRule2003], although there are also several important exceptions [@fanBergmannsRuleAllens2019; @stevensonAllensRuleNorth1986; @geistBergmannsRuleInvalid1987; @cuiFunctionalAdaptationRather2020]. Bergmann's rule is perhaps the most extensively studied macroecological rule to date, and several important reviews cover adherence to the rule [and lack thereof, @ashtonPatternsWithinSpeciesBody2002], potential mechanisms [@yom-tovRecentSpatialTemporal2011], implications in the face of climate change [@gardnerDecliningBodySize2011; @teplitskyClimateWarmingBergmanns2014], and best practices for empirical studies [@theriotHarnessingNaturalHistory2023]. Allen's rule is less studied [but see @rydingShapeshiftingChangingAnimal2021], despite evidence that it is equally important for thermoregulation [@frohlichAllometryRevealsTradeoffs2023].

Importantly, both rules stem from a central tenet: changes in body size or shape alter an organism’s surface-area-to-volume ratio, with direct consequences for thermoregulation. Metabolic scaling theory shows that heat production generally scales with body volume to the 3/4 power, while the area available for heat exchange scales with volume to the 2/3 power (assuming isometric growth, i.e., no change in shape). Thus, as size increases, heat production outpaces heat dissipation, unless the organism alters its shape to become proportionally longer (CITE Kleiber, Kleiber, Glazier).

Given their shared thermoregulatory basis, I propose a combined ecogeographical framework, the *Thermo-Morpho Rule*, which views size and shape as complementary axes of thermoregulatory adaptation. For scientists, the Thermo-Morpho rule implies that Bergmann's and Allen's rule should be examined together, given that similar changes in surface-area-to-volume could be achieved through changes in size, shape, or both (@fig-sa-to-v). In this perspective, I review literature published since 2021 to assess how often these rules have been analyzed jointly versus in isolation across birds and mammals, and what is lost when they are separated. I contend that a combined analysis is essential for a complete understanding of thermoregulatory adaptation under rapid climate change [@guoApplyingBergmannsRule2025]. Finally, I extend this logic beyond morphology, arguing that any phenotypic trait responsive to temperature (e.g., color, physiology, behavior)may interact with size and shape to influence thermal adaptation. I conclude by highlighting an integrated modeling framework that jointly estimates Bergmann’s and Allen’s rules and can be expanded to include additional traits, providing a practical path toward studying thermal adaptation across multiple traits.

![**Complementarity in Bergmann's and Allen's rules with increasing temperatures.** In response to increasing temperatures, organisms can achieve changes to their surface area to volume ratios by altering their a) body size, b) relative appendage size, c) or through complementarity, i.e. more minute changes to both body and appendage size. In c), morphological responses in body and appendage size are less pronounced than expected if Bergmann's and Allen's rules were operating in isolation. Figures were created by simulating changes to a cartoon bird made up of two spheres (representing the body and head) and a cone (representing the bill). Figure reproduced with permission from Baldwin et al (2023), see article for additional details.](../Figures/Baldwin_fig2.png){#fig-sa-to-v}

# Literature review

```{r}
#| label: Setup

# Load libraries
library(tidyverse)
library(readxl)
library(janitor)
library(smatr)
library(ggpubr)
library(cowplot)
ggplot2::theme_set(theme_cowplot())

# Source functions
source("Scripts/Key_allometry_fns.R")
```

```{r}
#| label: Load-in-format-data

Lr_allens <- read_excel("Lit_review/Methods_ecogeographic_rules_8.24.25.xlsx", 
                         sheet = "Updated Excel") %>% 
  clean_names() %>% 
  filter(include_exclude != "Exclude") %>%
  dplyr::select(authors, year, journal, bergmann) %>% 
  mutate(focus = ifelse(bergmann == "Y", "Both", "Allens_only")) %>% 
  dplyr::select(-bergmann)
Lr_bergs <- read_excel("Lit_review/Methods_ecogeographic_rules_8.24.25.xlsx", 
                         sheet = "Bergs studies") %>% 
  clean_names() %>% 
  dplyr::select(authors, year, journal, allens) %>% 
  mutate(focus = ifelse(allens == "Y", "Both", "Bergmanns_only")) %>% 
  dplyr::select(-allens)

Lr_ecogeo <- bind_rows(Lr_allens, Lr_bergs)

# Prep labeling tbl
Lr_p <- Lr_ecogeo %>% 
  summarize(num_studies = n(), .by = c(focus)) %>% 
  mutate(Tot_studies =  sum(num_studies), 
         Prop = num_studies / Tot_studies,
         Per = Prop * 100) %>% 
  mutate(Per = round(Per, 0),
         Prop = round(Prop, 2), 
         Per = paste0(Per, "%"),
         focus = str_replace(focus, "_", "\n"))
```

```{r}
#| label: fig-barplot-in-isolation
#| fig-cap: Bergmann's and Allen's rules are often examined in isolation. Of 48 studies published since 2021 examining Bergmann's rule, Allen's rule, or both, nearly half examined Bergmann's or Allen's rule in isolation. A focus on Bergmann's rule in isolation was much more common than focusing on Allen's rule in isolation, likely because Allen's rule is inherently multivariate. 
#| fig-height: 6.5

#Plot
Lr_ecogeo %>% 
  mutate(focus = str_replace(focus, "_", "\n")) %>% 
  ggplot() + 
  geom_bar(aes(x = focus, fill = factor(year)), 
           color = "black", stat = "count") + 
   geom_text(
    data = Lr_p, aes(x = focus, y = num_studies, label = Per),
    vjust = -0.7, size = 5
  ) + labs(x = NULL, y = "Frequency", fill = "Year")
```

```{r}
#| label: pull-percents

pull_per <- function(focus){
  Lr_p %>% filter(focus == {{ focus }}) %>% 
  pull(Per)
}
Per_allen <- pull_per(focus = "Allens\nonly")
Per_berg <- pull_per(focus = "Bergmanns\nonly")
Per_both <- pull_per(focus = "Both")
```

Of `r nrow(Lr_ecogeo)` studies published since 2021 examining Bergmann's rule, Allen's rule, or both, nearly half examined them in isolation. A focus on Bergmann's rule alone (`r Per_berg`) was far more common than on Allen's rule alone (`r Per_allen`), likely because Allen's rule studies generally measured *relative* appendage length (89% of studies), i.e., appendage length controlling for body size. Thus, Allen's rule is inherently a multivariate endeavor, and at some level, authors are already considering trends in body size across ecological gradients.

# The need for a combined approach

What do scientists risk by examining Bergmann's or Allen's rule in isolation? Beyond a loss of conceptual clarity, we may miss important shifts in surface area–to–volume ratios or reach incomplete (or even misleading) conclusions about wildlife responses to temperature gradients. Two recent global analyses assessing Bergmann's and Allen's rules in tandem highlight important coevolution in size and shape of birds [@baldwinComplementarityAllensBergmanns2023; @frohlichAllometryRevealsTradeoffs2023]. Both studies argue that the surface area–to–volume ratio is the key functional trait under selection, and that these rules need not operate independently or simultaneously to generate adaptive thermal responses.

@baldwinComplementarityAllensBergmanns2023 showed that small, coordinated shifts in both body size and appendage length can produce similar changes in surface area–to–volume ratios as larger shifts in either trait alone (@fig-sa-to-v). This “complementarity” was strongly supported as 81% of the 107 bird families examined did not conform to either rule when analyzed separately, and none conformed to both rules simultaneously. Instead, most families exhibited modest but coordinated changes in size and shape in the directions predicted by both rules. Such adjustments may buffer against the fitness costs of extreme changes in a single trait while still permitting meaningful shifts to thermoregulatory capacity.

@frohlichAllometryRevealsTradeoffs2023 also found evidence for coordinated evolution of size and shape in birds. However, instead of complementarity, they foundt hat species responded to warming either by decreasing body size or by increasing appendage length, but rarely both [also see @dubinerWidespreadRecentChanges2022]. @frohlichAllometryRevealsTradeoffs2023 interpreted this as support for the 'trade-off hypothesis', arguing that different thermoregulatory strategies carry distinct costs. For example, smaller body sizes can reduce competitive ability [@shipleySelectionCounteractsDevelopmental2022] or foraging efficiency [@vangilsBodyShrinkageDue2016], and changes to appendage size or shape can alter acoustic performance [@friedmanEvolutionMultifunctionalTrait2019; @sebastianelliContinentwidePatternsSong2022].

Importantly, the joint analytical approach (i.e., the joint analysis of Bergmann's and Allen's rule) these authors employed allowed a more nuanced interpretation of the morphology-environment relationship. For example, @baldwinComplementarityAllensBergmanns2023 showed that species that adhered to Bergmann's rule had less common bill shapes, likely because bird bills are highly functionally constrained by foraging [e.g., Trochilidae and Rallidae, @baldwinComplementarityAllensBergmanns2023]. Thus, these taxa relied on changes in body size to meet thermoregulatory needs. **The joint modeling** of Bergmann's and Allen's rules has also shown **differential morphological responsiveness across body sizes.** For example, smaller species tend to exhibit stronger morphological responses to warming than larger species [@mcqueenShorebirdsAreShrinking2024; @neate-cleggAfromontaneUnderstoryBirds2024], likely due to greater evolutionary potential (e.g., larger effective population sizes, higher genetic variance, and faster molecular evolution; Orkney & Hedrick 2024; @zimovaBodySizePredicts2023). Results from @frohlichAllometryRevealsTradeoffs2023 were particularly striking, as they found that morphology-temperature relationships often completely flipped sign when interactions between size and shape were included (@fig-interactions). For example, despite overall conformity to Bergmann’s rule, patterns were reversed (i.e., inverse Bergmann's rule) in birds with particularly short bills and long legs. Similarly with Allen's rule, the overall effect of temperature on relative appendage length was negative, but only in small-bodied birds (i.e., large-bodied birds conformed to Allen’s predictions). Conclusions would have been highly obfuscated without modeling the interactions between size, shape, and temperature, highlighting the need to include interactions when examining the codependency of ecogeographical rules [@gastonEcogeographicalRulesElements2008].

![**Modeling interactions between size and shape is critical in understanding tradeoffs in organismal response to increasing temperatures**. In predicting a species' experienced temperature envelope, Fröhlich et al (2023) found that the environment-morphology relationship of mass (top left), beak (top right), and tarsus (bottom) was dependent on the size and shape of the organism. All relationships were highly significant at p \< 0.001. See Figure 4 in Fröhlich et al (2023) for additional infromation; figure was modified and reproduced with permission from Arkadiusz Fröhlich.](../Figures/Frohlich_fig4F_modified.pdf){#fig-interactions}

Together, these studies underscore the need for joint analyses of size and shape to fully capture morphological responses to temperature gradients. Such models clarify the correlated evolutionary responses that underlie thermoregulation [@baldwinComplementarityAllensBergmanns2023] and help explain why isolated tests of Bergmann’s and Allen’s rules often yield mixed or contradictory results [@fanBergmannsRuleAllens2019; @stevensonAllensRuleNorth1986; @geistBergmannsRuleInvalid1987; @cuiFunctionalAdaptationRather2020]. Most species likely adjust their surface area–to–volume ratios across temperature gradients, but studies focusing on single (potentially constrained) traits may miss the adaptive signal entirely. Alternatively, species may exhibit complementarity — small, coordinated changes in both size and shape that escape detection when the rules are tested independently.

# Looking forward

## The Thermo-Morpho Rule

The Thermo-Morpho Rule views size and shape as complementary axes of thermoregulatory adaptation, providing theoretical justification for analyzing Bergmann’s and Allen’s rules in tandem. Adherence to this rule occurs when species follow Bergmann’s rule, Allen’s rule, or exhibit coordinated changes in body size and appendage length that together alter the surface area–to–volume ratio. This unified framework formalizes the idea that both size and shape contribute to thermoregulatory efficiency, and that evolution may act on either or both traits depending on functional constraints.

## Beyond morphology: Toward an integrative thermoregulatory framework

Thermoregulatory adaptation extends beyond morphology. Organisms can respond to thermal gradients through shifts in physiology, coloration, and behavior, suggesting that any trait under thermal selection could act in concert with morphology to fine-tune thermoregulation CITE Clarke, Rothery, & Isaac, 2010 dsd\[@tianPredictingBioticResponses2020\]. For instance, Bogert’s rule (the “thermal melanism hypothesis”) posits that darker individuals are favored in cooler environments because dark integuments absorb heat more efficiently \[Bogert 1949; Clusella-Trullas et al. 2007, 2008, 2009\]. Likewise, Hesse’s rule proposes that relative organ size, particularly the heart and lungs, increases in cooler climates to sustain higher metabolic demands \[Hesse 1924; Müller et al. 2014\], implying that warming climates could lead to smaller relative organ sizes as metabolic rates decline \[Seebacher 2009\].

Despite the potential for complementarity or trade-offs among ecogeographical rules, few studies have integrated them within a joint analytical framework [but see @cardiliniTestsEcogeographicalRelationships2016; @macedoMacroevolutionBodyExtremities2025]. For instance, @cuiFunctionalAdaptationRather2020 examined Bergmann's, Allen's, and Hesse's rules simultaneously and, consistent with the ‘trade-off hypothesis’, found support only for Allen’s rule, suggesting that body and heart size may be constrained by absolute energy efficiency. Finally, behavioral adjustments are also integral to thermoregulation. For example, appendage length moderated thermoregulatory behaviors, such that birds with longer bills and legs exposed them more to facilitate dry heat transfer. [Montmany et al. 2022]. Other behaviors like migration, torpor, hibernation, habitat selection and structure, and nocturnality likewise mediate the thermal niche and should continue to be investigated [@he2023global; @perez-amayaCenturylongShiftsBody2025] (CITE Bonebrake; Jirinec).

A natural next step is to formalize a broader *Thermo-Adaptive Rule*, encompassing how temperature jointly shapes morphology, physiology, coloration, and behavior. Such an approach will require data that are not yet routinely collected, but, as with morphological databases [@tobiasAVONETMorphologicalEcological2022], comparable datasets on physiological and color traits are rapidly emerging (e.g., [Project Plumage](https://www.nhm.ac.uk/take-part/monitor-and-encourage-nature/project-archive/project-plumage.html)). By integrating across these phenotypic and behavioral axes, we can move toward a more complete understanding of how organisms meet the thermoregulatory challenges of a warming world [@guoApplyingBergmannsRule2025].

## An integrated model

Moving forward, I recommend integrated models that jointly estimate Bergmann’s, Allen’s, and potentially other important ecogeographical rules [@fig-path-diagram, e.g., @frohlichAllometryRevealsTradeoffs2023]. For example, @macedoMacroevolutionBodyExtremities2025 compared hypotheses using phylogentically-controlled causal path analysis and found the strongest support for a model incorporating allometry alongside temperature's effects on body size (Bergmann's rule), appendage length (Allen's rule), and coloration of bare parts (Bogert’s rule).

Fortunately, integrated models can help resolve several additional issues that have plagued studies of Bergmann's and Allen's rules. First, body size is inherently multivariate, and no single trait can fully characterize an organism's structural size. Treating size as a latent variable means that the model can incorporate size information from an arbitrary number of traits. The second issue that this integrated model addresses is measurement error. With repeat measurements (at an appropriate time scale), each individual’s true (but latent) mass and length is reflected in the shared variance among their manifest indicators (M1–M3 and A1–A3 in @fig-path-diagram). Accounting for measurement error is particularly important for mass, which can vary tremendously due to physiological state (e.g., gut fill, defecation, reproductive or hydration), as well as length variables with low repeatability [e.g., those with small dimensions; @theriotHarnessingNaturalHistory2023; @subasingheRepeatabilityValidityPhenotypic2021]. Finally, the model accounts for allometric scaling in the calculation of relative appendage length [@symondsItsAllRelative2025], without assuming directional causality, i.e., that mass *causes* appendage length [@peigNewPerspectivesEstimating2009; @peigParadigmBodyCondition2010].

![Path diagram showing a hypothetical model that jointly tests Allen's and Bergmann's rule. Ovals represent latent variables and rectangles represent measured variables. M1 - M3 represent multiple measurements of mass at an appropriate time scale; A1 - A3 represent three distinct appendages (e.g., tarsus, wing, and bill length). Intercepts and residual errors are implied and thus not shown. Importantly, this framework could easily be extended to test additional hypotheses related to physiology, coloration, or behavior.](../Figures/path_diagram.pdf){#fig-path-diagram}

# Conclusions

Substantial resources go into collecting and analyzing phenotypic data, particularly morphology. Yet the Anthropocene demands fast and informed decision-making, conservationists cannot afford to split hairs over legacy ecogeographical rules. The hyperfocus on classifying trait variation into separate rules is ultimately unhelpful [@goldenbergLinkBodySize2022]. Instead, analyzing multiple traits simultaneously within a holistic framework offers greater power to understand how entire phenotypes respond to temperature and other environmental gradients. These considerations apply across both intra- and inter- specific scales. Intraspecific analyses can forecast how species will respond to rising temperatures within their ranges and identify vulnerable populations, whereas interspecific analyses can reveal how entire clades respond and highlight groups of organisms particularly susceptible to warming. Ultimately, identifying the phenotypic pathways — whether shifts in coloration, body or organ size, or appendage length — through which selection operates will enable conservationists to anticipate biological responses in an increasingly warm and human-dominated world. The Thermo-Adaptive Rule provides a theoretical foundation and an urgent call to integrate multiple phenotypic axes under a common framework, deepening our understanding of environment–phenotype relationships and sharpening predictions of biological responses to a warming world.

# Data availability

All studies included in the literature review are available for download at LINK, and the full list of citations are available in the Supporting Information.

# Acknowledgements

Thank you to Justin Baldwin and Arkadiusz Fröhlich for letting me use their excellent conceptual figures.

# Letter to editor

Nature comms - 'avoid repeating information that is already present in the abstract and introduction'

Given this shared thermoregulatory basis, some authors have discussed the interrelatedness or shared mechanistic basis of macrogeographic rules, but none have explicitly argued for their joint and integrated analysis.

Authors have advocated for the integration of macroecological rules across broad trait 'categories' (e.g., morphology, physiology, life history, population dynamics; Ashton 2001, Blackburn & Gaston 2001, Gaston, Chown, and Evans 2008), yet there has been no similar perspective arguing for this integration within morphological traits, and even less a review assessing the state of the literature.

Given the relatively large body of literature examining size and shape in tandem I focus this perspective piece there before moving to a broader focus on phenotypic response. I propose two new integrated rules: the Thermo-morpho rule and the Thermo-Adaptive rule , corresponding to temperature's impact on size and shape and the broader phenotype, respectively. Ultimately, I argue that scientists should strive to test the Thermo-Adaptive rule, but due to logistical constraints and data availability issues, I do not expect this will be commonplace for a number of years. Thus, even just combining the analyses of size and shape (i.e., the Thermo-morpho rule) would be a huge improvement over the still-to-common siloed treatment of Bergmann's and Allen's rules.

This is critical and is recommended in some empirical studies, but no clear synthesis exists

In the 'An integrated model' section I note that this model addresses 3 longstanding issues in studies of environment-morphology studies: 1) body size is inherently multivariate, 2) measurement error, and 3) accounting for allometric scaling without assuming directional causality. I do believe these are extremely important issues, but this may be a topic for a different paper. I am happy to remove the 'An integrated model' section altogether if reviewers see it as out of scope with the rest of the paper.

Format - Review studies since @rydingShapeshiftingChangingAnimal2021 important review synthesizing shape-shifting in the face of climate change, don't need figure / to cite all the articles , could be included in Supp information

# Extras - ignore

## To do

-   Short methods section for literature review (Supp info?)

-   Letter to the editors

-   Citation style?

-   Ashton 2001 & Gaston, Chown, and Evans 2008 - few authors have called for a joint analysis of ecogeographical patterns (but see Ashton 2001). Very briefly, literally in the last paragraph, discusses the interrelatedness between four rules. Gaston et al (2008) paper discusses idea of shared mechanism more. None of them go beyond a brief mention of Allen's rule

-   Mention importance of habitat (Greenberg et al, 2012) & habitat fragmentation (Desrochers 2010, Jirinec 2021, Hermes 2016, maybe Ausprey 2023)?

-   Mention migration (Dubiner & Meiri 2022; McQueen 2024), behavior, 'activity patterns' (see Cunnigham, or originally Symonds 2025), tropical vs temperate (Ryding et al, 2025), other functional constraints (e.g. diet), see He et al (2023). Guo says a better understanding of the alternative drivers is helpful for conservation.\
    Cunningham SJ, Gardner JL, Martin RO (2021) Opportunity costs and the response of birds and mammals to climate warming. Front Ecol Env 19:300–307. <https://doi.org/10.1002/fee.2324>

    -   See Montmany et al (2022) may be relevant for behavior - 'Body postural adjustments, panting, and locomotor activity' were the response variable and relative appendage lengths were the predictors (the study did not assess Allen's rule)

    -   Migratory species size and shape is constrained as they need 'to carry and conserve energy during long-­distance travel'. Also they avoid the harshest climates through migration

    -   Migration easy to test interspectific (AVONET) and intraspecific with the increase of tracking data (e.g., Skinner et al 2025; Shipley I am pretty sure, cited in my own paper)

    -   'local environment conditions, such as temperature variability, resource availability, or habitat structure, likely mediate morphological responses to climate warming' Ocampo-Peñuela (2024), quote from Perez-Amaya (2025)

    -   Perez-Amaya (2025) - Habitat structure

-   Interactions - McQueen 2024 tested models with interactions between year × mass, year × population location, and year × migration behaviour

    -   Instead of interactions, Neate-clegg (afromontane birds) uses the full bayesian posteriors to model rates of body mass change as a function of body size

-   Include precipitation in models too as precipitation is also changing rapidly under climate change [@macedoMacroevolutionBodyExtremities2025], (Benham & Bowie, 2021 - The influence of spatially heterogeneous anthropogenic change on bill size evolution in a coastal songbird)

## Simulation

```{r}
#| label: Sim-bergs_rule_conflicting

Ex_df_hypo <- pmap(tibble(b_avg_12 = .22, r_12 = .3, r_13 = -.4, r_23 = -.1), gen_data)[[1]]

# NOTE:: In paper make note that we would know this will happen due to hypoallometry

Ex_df_hypo2 <- Ex_df_hypo %>% mutate(SA_V = Append^2 / Mass) #%>% 
  #filter(Append > 0 & Mass > 0 & SA_V < 5)
```

```{r}
#| label: Confirm-sma-slope
#| include: false

smatr::sma(Append ~ Mass, data = Ex_df_hypo2)
ref_slope <- 0.33
ref_intercept <- mean(Ex_df_hypo2$Append) - ref_slope * mean(Ex_df_hypo2$Mass)

p4 <- ggplot(Ex_df_hypo2, aes(x = Mass, y = Append)) +
  geom_point(alpha = .4) + 
  ggpmisc::stat_ma_line(method = "SMA", color = "blue") +
  geom_abline(slope = 0.33, intercept = ref_intercept, 
              linetype = "dashed", color = "red")
```

```{r}
#| label: fig-bergs_rule_conflicting

p1 <- ggplot(Ex_df_hypo2, aes(x = Temp_inc, y = Append)) +
  geom_point(alpha = .4) + 
  geom_smooth(method = "lm", se = FALSE) + 
  labs(x = "Temperature increase", y = "Appendage") 

p2 <- ggplot(Ex_df_hypo2, aes(x = Temp_inc, y = Mass)) +
  geom_point(alpha = .4) + 
  geom_smooth(method = "lm", se = FALSE) + 
  labs(x = "Temperature increase", y = "Mass")

# We would expect that SA:V would increase as temperature increases, but in this case SA:V decreases
p3 <- ggplot(Ex_df_hypo2, aes(x = Temp_inc, y = SA_V)) +
  geom_point(alpha = .4) + 
  geom_smooth(method = "lm", se = FALSE) + 
  labs(y = "Surface area to volume", x = "Temperature increase")

ggarrange(p4, p1, p2, p3, nrow = 2, ncol = 2, labels = 'AUTO')
```

Understanding the allometric scaling relationship for a species can aid in interpretation of results. For example, imagine a study using two variables as metrics of body size (wing chord and mass), and a third variable $\text{wing length}^2$ divided by body mass as a more direct metric of the surface area to volume ratio [e.g. @nwaoguTemperatureAridityDetermine2018]. The authors are surprised as the hypothetical species appears to adhere to Bergmann's rule when mass and wing chord are used separately as the response variables, but they find a result contradicting Bergmann's rule when using $\text{wing length}^2$ divided by body mass as the response variable.

## General

The inclusion of interactions between temperature, size, and shape can reveal the co-evolution of thermoregulatory strategies in wildlife.

'Human modifications of mammal species distributions lead to substantially underestimating the strength of Bergmann’s rule (Faurby, 2016).'

There is substantial evidence documenting morphological change in response to changing climate. But what do conservation measures look like in light of our understanding of phenotypic changes? Most studies do not dare make concrete recommendations of how this might influence management or conservation planning.

Complementarity (ROODENRIJS et al black oystercatchers, 2024, SFU group)

Finally, when predicting the temperatures experienced across species’ ranges, the strongest model supported a three-way interaction between body size, tarsus length, and beak length (see @fig-interactions). Formulating the experienced climate envelope as the response variable was an especially clever way to reveal the multidimensional environment-morphology relationship.

Integrated model:

1\) Null allometric model: Mass' influence on absolute appendage length

2\) Bergmann’s rule model: The effects included in the null allometric model + temperature's influence on mass

3\) Allen’s rule model: The effects included in the null allometric model + temperature's influence on appendage length

4\) ‘Mixed’ model: The combination of models 1-3 (Allometry + Bergmann's + Allen's rules)

Still, these results were somewhat surprising given that larger birds are predicted to be more sensitive to heat stress given lower SA:V ratios (Geiser and Turbill 2009; McCain and King 2014).

Indeed, models with extensive interactions revealed codependence between size and shape in the evolutionary struggle for thermoregulation.

Indeed, @baldwinComplementarityAllensBergmanns2023 showed there was a tight negative relationship between changes in body size and appendage length, with larger shifts in one trait associated with smaller shifts in the other.

{{< pagebreak >}}

# References
