---
title: "Beyond Bergmann and Allen: A Unified Framework for Thermoregulatory Morphology"
number-sections: true
# If you do not want code to show... echo: false
execute: 
 echo: false
 message: false
 warning: false
format: pdf
editor: visual
bibliography: ../Supp_files/Morphology.bib
link-citations: true
csl: ../Supp_files/methods-in-ecology-and-evolution.csl
knitr:
  opts_chunk: 
    comment: "#>"
    collapse: true
editor_options: 
  chunk_output_type: console
---

{{< pagebreak >}}

# Abstract

Understanding how animals’ size and shape vary with climate is central to predicting their responses to global change. Yet scientists often study these traits separately, even though both influence heat exchange with the environment. I propose a unifying ecogeographic rule which treats size and shape as complementary strategies for thermoregulation. Finally, I highlight the need for a broader thermo-adaptive framework, given that organisms can respond to increasing temperatures via shifts in physiology, behavior, or other properties of the fur or feathers. Recognizing the interdependence of thermoregulatory mechanisms is critical in an era of rapid global change.

# Introduction

We are in the Anthropocene — a period of rapid global change dominated by habitat loss and climate change. While many species have not kept pace with these changes and have gone extinct, other species have been able to alter their shapes, sizes, ranges, and behaviors to adjust to this new human-dominated world [@parmesanEcologicalEvolutionaryResponses2006; @gardnerDecliningBodySize2011; @millienEcotypicVariationContext2006]. Understanding the sub-lethal, plastic or evolutionary responses is critical to understanding humanities’ full impact on biodiversity [@guoApplyingBergmannsRule2025]. With this knowledge, we will be better able to forecast changes to the natural world, as well as their impacts on ecosystem service provisioning [@luckImprovingApplicationVertebrate2012].

An organism’s morphology is perhaps the most fundamental determinant of how it interacts with its environment, and thus there are strong selective pressures to ensure that morphology is well-suited to its environment [@pigotMacroevolutionaryConvergenceConnects2020]. Thus, scientists have long sought to understand how morphology is shaped by the environment [e.g., @winklerEcomorphologyMigrants1992], and some of the oldest macrogeographical 'rules' attempt to make sense of patterns in body size and shape. For example, Bergmann’s rule posits that individuals or populations of endothermic species tend to be larger in colder climates, presumably to minimize heat loss via a reduced surface-area-to-volume ratio [@bergmann1848ueber]. The complementary Allen’s rule states that appendages (e.g., limbs, tails, bills) tend to be shorter in colder regions, again reducing heat loss via a decreased surface-area-to-volume ratio [@allen1877influence].

There is general support for both Bergmann’s and Allen's rules in endotherms [@teplitskyClimateWarmingBergmanns2014; @rydingShapeshiftingChangingAnimal2021; @meiriValidityBergmannsRule2003], although there are also several important exceptions [@fanBergmannsRuleAllens2019; @stevensonAllensRuleNorth1986; @geistBergmannsRuleInvalid1987; @cuiFunctionalAdaptationRather2020]. Bergmann's rule is perhaps the most extensively studied macroecological rule to date, and several important reviews cover adherence to the rule [and lack thereof, @ashtonPatternsWithinSpeciesBody2002], potential mechanisms [@yom-tovRecentSpatialTemporal2011], implications in the face of climate change [@gardnerDecliningBodySize2011; @teplitskyClimateWarmingBergmanns2014], and best practices for empirical studies [@theriotHarnessingNaturalHistory2023]. Allen's rule is less studied [but see @rydingShapeshiftingChangingAnimal2021], despite evidence that it is equally important for thermoregulation [@frohlichAllometryRevealsTradeoffs2023].

Importantly, both rules stem from a central tenet: changes in body size or shape alter an organism’s surface-area-to-volume ratio, with direct consequences for thermoregulation. Metabolic scaling theory shows that heat production generally scales with body volume to the 3/4 power, while the area available for heat exchange scales with volume to the 2/3 power (assuming isometric growth, i.e., no change in shape). Thus, as size increases, heat production outpaces heat dissipation, unless the organism alters its shape to become proportionally longer [@kleiber1932body; @kleiber1972body; @glazier2008effects].

Given their shared thermoregulatory basis, I propose a combined ecogeographical framework, the *Thermo-Morpho Rule*, which views size and shape as complementary axes of thermoregulatory adaptation. For scientists, the Thermo-Morpho rule implies that Bergmann's and Allen's rule should be examined together, given that similar changes in surface-area-to-volume could be achieved through changes in size, shape, or both (@fig-sa-to-v). In this perspective, I review literature published since 2021 to assess how often these rules have been analyzed jointly versus in isolation across birds and mammals, and what is lost when they are separated. I contend that a combined analysis is essential for a complete understanding of thermoregulatory adaptation under rapid climate change [@guoApplyingBergmannsRule2025]. Finally, I extend this logic beyond morphology, arguing that any phenotypic trait responsive to temperature (e.g., color, physiology, behavior)may interact with size and shape to influence thermal adaptation. I conclude by highlighting an integrated modeling framework that jointly estimates Bergmann’s and Allen’s rules and can be expanded to include additional traits, providing a practical path toward studying thermal adaptation across multiple traits.

![**Complementarity in Bergmann's and Allen's rules with increasing temperatures.** In response to increasing temperatures, organisms can achieve changes to their surface area to volume ratios by altering their a) body size, b) relative appendage size, c) or through complementarity, i.e. more minute changes to both body and appendage size. In c), morphological responses in body and appendage size are less pronounced than expected if Bergmann's and Allen's rules were operating in isolation. Figures were created by simulating changes to a cartoon bird made up of two spheres (representing the body and head) and a cone (representing the bill). Figure reproduced with permission from Baldwin et al (2023), see article for additional details.](../Figures/Baldwin_fig2.png){#fig-sa-to-v}

# Literature review

```{r}
#| label: Setup

# Load libraries
library(tidyverse)
library(readxl)
library(janitor)
library(smatr)
library(ggpubr)
library(cowplot)
ggplot2::theme_set(theme_cowplot())

# Source functions
source("Scripts/Key_allometry_fns.R")
```

```{r}
#| label: Load-in-format-data

Lr_allens <- read_excel("Lit_review/Methods_ecogeographic_rules_8.24.25.xlsx", 
                         sheet = "Allens") %>% 
  clean_names() %>% 
  filter(include_exclude != "Exclude") %>%
  dplyr::select(authors, year, journal, bergmann) %>% 
  mutate(focus = ifelse(bergmann == "Y", "Both", "Allens_only")) %>% 
  dplyr::select(-bergmann)
Lr_bergs <- read_excel("Lit_review/Methods_ecogeographic_rules_8.24.25.xlsx", 
                         sheet = "Bergmanns") %>% 
  clean_names() %>% 
  filter(include_exclude != "Exclude") %>%
  dplyr::select(authors, year, journal, allens) %>% 
  mutate(focus = ifelse(allens == "Y", "Both", "Bergmanns_only")) %>% 
  dplyr::select(-allens)

Lr_ecogeo <- bind_rows(Lr_allens, Lr_bergs)

# Prep labeling tbl
Lr_p <- Lr_ecogeo %>% 
  summarize(num_studies = n(), .by = c(focus)) %>% 
  mutate(Tot_studies =  sum(num_studies), 
         Prop = num_studies / Tot_studies,
         Per = Prop * 100) %>% 
  mutate(Per = round(Per, 0),
         Prop = round(Prop, 2), 
         Per = paste0(Per, "%"),
         focus = str_replace(focus, "_", "\n"))
Lr_p 
```

```{r}
#| label: pull-percents

pull_per <- function(focus){
  Lr_p %>% filter(focus == {{ focus }}) %>% 
  pull(Per)
}
Per_allen <- pull_per(focus = "Allens\nonly")
Per_berg <- pull_per(focus = "Bergmanns\nonly")
Per_both <- pull_per(focus = "Both")
```

I conducted a literature review of empirical studies published in the past 5 years (2021-2025) to determine how often Bergmann's and Allen's rules are examined together versus in isolation. I queried the Web of Science to identify empirical studies testing Bergmann's or Allen's rule in extant species of birds and mammals that examined at least one continuous morphological variable in response to an environmental gradient (e.g., temperature, precipitation) or a proxy (e.g., latitude, elevation, year). The search was conducted on November 4th, 2025, using the following search terms: (TS=(Allen\* OR Bergmann\* OR "morpholog\* change" OR limb\* OR appendage*) AND TS=(bird* OR avian OR aves OR mammal\* OR endotherm\* OR homeotherm*) AND TS=(latitud* OR temperature OR year OR thermoregulat\* OR climat\* OR elevation OR "environmental gradient" OR temporal*)) NOT TS=(fossil* OR paleontolog\* OR Cretaceous OR Jurassic OR Miocene OR Pleistocene) AND DT=(ARTICLE OR "EARLY ACCESS") AND WC=(ECOLOGY OR ZOOLOGY OR "EVOLUTIONARY BIOLOGY" OR BIOLOGY OR "VETERINARY SCIENCES" OR "BIODIVERSITY CONSERVATION" OR ORNITHOLOGY OR "ANATOMY MORPHOLOGY" OR PHYSIOLOGY). The initial query produced 193 studies which were manually reviewed to ensure conformity to the inclusion criteria, of which XX were eliminated. In coding the remaining YY articles, I found an additional ZZ studies that met the criteria for inclusion but had not been identified by the original query, yielding a final corpus of TOTAL \# studies used in the current review.

Of `r nrow(Lr_ecogeo)` studies examining Bergmann's or Allen's rules, nearly half examined them in isolation. A focus on Bergmann's rule alone (`r Per_berg`) was far more common than on Allen's rule alone (`r Per_allen`), likely because Allen's rule studies generally measured *relative* appendage length (89% of studies), i.e., appendage length controlling for body size. Thus, Allen's rule is inherently a multivariate endeavor, and at some level, authors are already considering trends in body size across ecological gradients.

### Data availability

My coding of the TOTAL studies is available for download in the supporting information. The full set of articles can be examined by reproducing the Web of Science search.

# The need for a combined approach

What do scientists risk by examining Bergmann's or Allen's rule in isolation? Beyond a loss of conceptual clarity, I argue that we may miss 1) trade-offs - where the studied trait (e.g., body size) is functionally constrained and thus doesn't respond to temperature; 2) complementarity -combined changes to size and shape that together amount to important adjustments to the surface area–to–volume ratio, and finally, 3) interactions - codependency between size and shape in response to temperature gradients. Two recent global analyses assessing Bergmann's and Allen's rules in tandem highlight important coevolution in size and shape of birds [@baldwinComplementarityAllensBergmanns2023; @frohlichAllometryRevealsTradeoffs2023]. Both studies argue that the surface area–to–volume ratio is the key functional trait under selection, and that these rules need not operate independently or simultaneously to generate adaptive thermal responses.

Most species likely adjust their surface area–to–volume ratios across temperature gradients, but studies focusing on Bergmann's or Allen's rules in isolation may miss the adaptive signal entirely. Studies with large taxonomic scope have found that species responded to warming either by decreasing body size or by increasing appendage length, but rarely both [@frohlichAllometryRevealsTradeoffs2023; @dubinerWidespreadRecentChanges2022]. @frohlichAllometryRevealsTradeoffs2023 interpreted this as support for the 'trade-off hypothesis', arguing that different thermoregulatory strategies carry distinct costs. For example, smaller body sizes can reduce competitive ability [@shipleySelectionCounteractsDevelopmental2022] or foraging efficiency [@vangilsBodyShrinkageDue2016], and changes to appendage size or shape can alter acoustic performance [@friedmanEvolutionMultifunctionalTrait2019; @sebastianelliContinentwidePatternsSong2022]. Thus, life history can constrain thermoregulatory adaptation. For example, taxa that relied on changes in body size to meet thermoregulatory needs had less common bill shapes due to foraging constraints [e.g., Trochilidae and Rallidae, @baldwinComplementarityAllensBergmanns2023]. This is also an important reminder of the value of using multiple traits as a proxy for body size, given that they can be under distinct selective pressures [@baileyUsingDifferentBody2020].

@baldwinComplementarityAllensBergmanns2023 showed that small, coordinated shifts in both body size and appendage length can produce similar changes in surface area–to–volume ratios as larger shifts in either trait alone (@fig-sa-to-v). This “complementarity” was strongly supported as 81% of the 107 bird families examined did not conform to either rule when analyzed separately, and none conformed to both rules simultaneously. Instead, most families exhibited modest but coordinated changes in size and shape in the directions predicted by both rules [also see @dubinerWidespreadRecentChanges2022]. Such adjustments may buffer against the fitness costs of extreme changes in a single trait while still permitting meaningful shifts to thermoregulatory capacity.

Analytically, the joint examination of Bergmann's and Allen's rule can allow for a more nuanced interpretation of morphology-environment relationships, particularly by testing interactions between size, shape, and temperature. For example, **recent work shows** that smaller species tend to exhibit stronger morphological responses to warming than larger species [@mcqueenShorebirdsAreShrinking2024; @neate-cleggAfromontaneUnderstoryBirds2024], likely due to greater evolutionary potential (e.g., larger effective population sizes, higher genetic variance, and faster molecular evolution; Orkney & Hedrick 2024; @zimovaBodySizePredicts2023). Similarly, @frohlichAllometryRevealsTradeoffs2023 showed that size, shape, and their interaction were important for understanding morphology-temperature relationships. For example, despite overall conformity to Bergmann’s rule, patterns were reversed (i.e., inverse Bergmann's rule) in birds with particularly short bills and long legs. Similarly with Allen's rule, the overall effect of temperature on relative appendage length was negative, but only in small-bodied birds (i.e., large-bodied birds conformed to Allen’s predictions). This highlights the codependency of ecogeographical rules, as relationships regularly flipped sign when interactions between size and shape were included [@fig-interactions, @gastonEcogeographicalRulesElements2008].

![**Modeling interactions between size and shape is critical in understanding tradeoffs in organismal response to increasing temperatures**. In predicting a species' experienced temperature envelope, Fröhlich et al (2023) found that the environment-morphology relationship of mass (top left), beak (top right), and tarsus (bottom) was dependent on the size and shape of the organism. All relationships were highly significant at p \< 0.001. See Figure 4 in Fröhlich et al (2023) for additional infromation; figure was modified and reproduced with permission from Arkadiusz Fröhlich.](../Figures/Frohlich_fig4F_modified.pdf){#fig-interactions}

# Looking forward

## The Thermo-Morpho Rule

The Thermo-Morpho Rule views size and shape as complementary axes of thermoregulatory adaptation, providing theoretical justification for analyzing Bergmann’s and Allen’s rules in tandem. Adherence to this rule occurs when species follow Bergmann’s rule, Allen’s rule, or exhibit coordinated changes in body size and appendage length that together alter the surface area–to–volume ratio. This unified framework formalizes the idea that both size and shape contribute to thermoregulatory efficiency, and that evolution may act on either or both traits depending on functional constraints [@frohlichAllometryRevealsTradeoffs2023; @baldwinComplementarityAllensBergmanns2023]. The lack of studies incorporating the Thermo-Morpho rule helps explain why Bergmann’s and Allen’s rules have yielded mixed or contradictory results [@fanBergmannsRuleAllens2019; @stevensonAllensRuleNorth1986; @geistBergmannsRuleInvalid1987; @cuiFunctionalAdaptationRather2020].

## An integrated model

Moving forward, I recommend integrated models that jointly estimate Bergmann’s, Allen’s, and potentially other important ecogeographical rules [@fig-path-diagram, e.g., @frohlichAllometryRevealsTradeoffs2023]. For example, @macedoMacroevolutionBodyExtremities2025 compared hypotheses using phylogentically-controlled causal path analysis and found the strongest support for a model incorporating allometry alongside temperature's effects on body size (Bergmann's rule), appendage length (Allen's rule), and coloration of bare parts (Bogert’s rule).

![Path diagram showing a hypothetical model that jointly tests Allen's and Bergmann's rule. Ovals represent latent variables and rectangles represent measured variables. M1 - M3 represent multiple measurements of mass at an appropriate time scale; A1 - A3 represent three distinct appendages (e.g., tarsus, wing, and bill length). Intercepts and residual errors are implied and thus not shown. Importantly, this framework could easily be extended to test additional hypotheses related to physiology, coloration, or behavior.](../Figures/path_diagram.pdf){#fig-path-diagram}

## Beyond morphology: Toward an integrative thermoregulatory framework

Thermoregulation extends beyond morphology, as organisms can respond to increasing temperatures through shifts in physiology, behavior, phenology, or other properties of the fur or feathers [e.g. coloration, @millienEcotypicVariationContext2006; @mayrGeographicalCharacterGradients1956]. For instance, Bogert’s rule (the “thermal melanism hypothesis”) posits that darker individuals are favored in cooler environments because dark integuments absorb heat more efficiently [@bogertThermoregulationReptilesFactor1949; @clusella-trullasTestingThermalMelanism2008]. Likewise, Hesse’s rule proposes that relative organ size, particularly the heart and lungs, increases in cooler climates to sustain higher metabolic demands [@mullerRelativeHeartSize2014], implying that warming climates could lead to smaller relative organ sizes as metabolic rates decline [@seebacherResponsesTemperatureVariation2009].

The potential for trade-offs, complementarity, and interactions is not restricted to Bergmann's and Allen's rules, but is likely applicable to any of the adaptive thermoregulatory approaches mentioned above. For instance, woodrats (*Neotoma*) lack specialized adaptations to extreme heat despite living in desert environments, thus, their body size is extremely sensitive to temperature gradients [@brownBergmannsRuleClimatic1969; @smithPredictingWoodratNeotoma2006; [@millienEcotypicVariationContext2006]. This implies a trade-off between body size evolution and physiological adaptations common to desert species. Similarly, @mollerMorphologicalConstraintsChanging2017 showed that birds with high aspect ratios were unable to adapt their migration phenology as readily as those with low aspect ratios, suggesting a trade-off between morphological and phenological adaptation mechanisms. Empirical and simulated work has shown that size and the properties of pelage (e.g., fur depth) act complementarily to meet thermoregulatory needs in mammals [@briscoeBergmannMeetsScholander2015; @reynoldsEffectsBodySize1993]. Finally, several studies have documented interactions between morphology and behavior. Indeed, the relative appendage length of shorebirds moderated thermoregulatory behaviors, such that those with longer bills and legs exposed them to the environment more often to facilitate dry heat transfer [@pavlovicMorphologyGeographyPredict2019; @ryelandBillSizeMediates2017; @ryelandLegLengthTemperature2019; @playa-montmanyBehaviouralPhysiologicalResponses2023]. Other behaviors like migration, torpor, hibernation, habitat selection and structure, and nocturnality could also mediate the thermal environment experienced by wildlife \[@he2023global; @brivioSeekingTemporalRefugia2024; @jirinecLightThermalNiches2022; @bonebrakeClimateChangeThermoregulatory2020\]. These studies suggest that any trait under thermoregulatory selection could act in concert with, or in place of, morphological adaptation [@clarkeScalingBasalMetabolic2010; @tianPredictingBioticResponses2020].

A natural next step is to formalize a broader *Thermo-Adaptive Framework*, encompassing how thermal gradients jointly shape morphology, physiology, coloration, and behavior. The Thermo-Adaptive Framework is best viewed as a conceptual scaffold linking thermoregulatory function across traits. It is not an explicit rule that can be tested in the way that the Thermo-Morpho rule could be, but instead its purpose to raise awareness of trade-offs, complementarity, and interactions amongst adaptive thermoregulatory approaches. Such an approach will require data that are not yet routinely collected, but, as with morphological databases [@tobiasAVONETMorphologicalEcological2022], comparable datasets on physiological and color traits are rapidly emerging (e.g., [Project Plumage](https://www.nhm.ac.uk/take-part/monitor-and-encourage-nature/project-archive/project-plumage.html)). By integrating across these phenotypic and behavioral axes, we can move toward a more complete understanding of how organisms meet the thermoregulatory challenges of a warming world [@guoApplyingBergmannsRule2025].

# Conclusions

Substantial resources go into collecting and analyzing phenotypic data to determine adherence to ecogeographical rules. In the Anthropocene where conservationists must make fast and informed decisions, the hyperfocus on classifying trait variation into separate rules is ultimately unhelpful [@goldenbergLinkBodySize2022]. Instead, analyzing multiple traits simultaneously within a holistic framework offers greater power to understand how entire phenotypes respond to temperature and other environmental gradients. These considerations apply across both intra- and inter- specific scales. Intraspecific analyses can forecast how species will respond to rising temperatures within their ranges and identify vulnerable populations, whereas interspecific analyses can reveal how entire clades respond and highlight groups of organisms particularly susceptible to warming. Ultimately, identifying the phenotypic pathways — whether shifts in morphology, physiology, behavior, or coloration — through which selection operates will enable conservationists to anticipate biological responses in an increasingly warm and human-dominated world. I end with three concrete recommendations for ecographers, conservationists, and evolutionary biologists: 1) stop testing Bergmann's and Allen's rules separately; 2) increase awareness of trade-offs, complementarity, and interactions; and 3) work towards a broader Thermo-Adaptive framework to sharpen predictions of biological responses to a warming world.

# Acknowledgements

Thank you to Justin Baldwin and Arkadiusz Fröhlich for permitting the use of their excellent conceptual figures. Thank you to M Lynne Cooper for her help and advice throughout the review process.

# Extras - ignore

## To do

-   Several authors have recognized the competing causes that could influence body size (e.g., competition), appendage length (e.g., foraging), coloration, etc. (see flow chart in Yom-Tov & Geffen, 2011), but there has been less focus on how environmental conditions influence several different traits.

    -   If the focus is about body size then yes, considering the factors influencing body size is important. If study is about thermoregulation / climate change, then considering the alternative approaches for thermoregulation is important

    -   In formalizing the Thermo-adaptive rule, what is important for conservation? Increasing movement away from cities (urban heat islands) might be, but morphology isn't? (What can humans control?)

-   Millien et al (2006) paper has some important big picture stuff

    -   3 main ways that organisms respond to climate

    -   'Three main patterns of geographical phenotypic differentiation between populations within a species: (i) variation in colour; (ii) variation in the size of appendages; and (iii) variation in body size (Allen 1877).'

    -   'hair-rule (mammals have longer and thicker hairs in colder areas)'

        -   Briscoe et al (2015) Koala paper fur, Scholander 1955; STEUDEL, K., PORTER, W.P., and SHER, D. 1994

    -   'Moreover, there may be several evolutionary solutions for a given species to adapt to its new environment (Mayr 1956).'

    -   'Phenotypical and phenological changes occur at the smaller time scales, from a few years up to several 100 years. Shifts in species distributions and extinctions are detectable at larger time scales.'

-   Short methods section for literature review (Supp info?)

-   Shorten abstract

    -   Example (149 words): Understanding how organismal size and shape vary across environmental gradients is central to ecology, evolution, and conservation. Yet these traits are often studied separately, despite sharing a common thermoregulatory basis. Nearly half of 48 studies published since 2021 examined Bergmann’s or Allen’s rule in isolation, underscoring the need for a unified approach. I propose the *Thermo-Adaptive Rule*, which views size and shape as complementary axes of thermoregulation, both governed by surface area–to–volume relationships that mediate heat exchange. Recent joint analyses of Bergmann’s and Allen’s rules reveal that species respond to warming either by shrinking in size, elongating appendages, or showing subtle coordinated shifts that together alter thermal efficiency. Extending this framework to include physiological, behavioral, and color traits would further clarify the mechanisms underlying thermal adaptation. Integrative models that embrace multiple trait dimensions will be essential for predicting organismal responses to rapid global change.

-   Citation style?

-   Ashton 2001 & Gaston, Chown, and Evans 2008 - few authors have called for a joint analysis of ecogeographical patterns (but see Ashton 2001). Very briefly, literally in the last paragraph, discusses the interrelatedness between four rules. Gaston et al (2008) paper discusses idea of shared mechanism more. None of them go beyond a brief mention of Allen's rule

-   Mention importance of habitat (Greenberg et al, 2012) & habitat fragmentation (Desrochers 2010, Jirinec 2021, Hermes 2016, maybe Ausprey 2023)?

-   Mention migration (Dubiner & Meiri 2022; McQueen 2024), behavior, 'activity patterns' (see Cunnigham, or originally Symonds 2025), tropical vs temperate (Ryding et al, 2025), other functional constraints (e.g. diet), see He et al (2023). Guo says a better understanding of the alternative drivers is helpful for conservation.\
    Cunningham SJ, Gardner JL, Martin RO (2021) Opportunity costs and the response of birds and mammals to climate warming. Front Ecol Env 19:300–307. <https://doi.org/10.1002/fee.2324>

    -   Migratory species size and shape is constrained as they need 'to carry and conserve energy during long-­distance travel'. Also they avoid the harshest climates through migration

    -   Migration easy to test interspectific (AVONET) and intraspecific with the increase of tracking data (e.g., Skinner et al 2025; Shipley I am pretty sure, cited in my own paper)

    -   'local environment conditions, such as temperature variability, resource availability, or habitat structure, likely mediate morphological responses to climate warming' Ocampo-Peñuela (2024), quote from Perez-Amaya (2025)

        -   The context-dependency of the two rules: Others have found results generally consistent with Bergmann's rule, but found that smaller birds with increased wingyness can be selected for at higher elevations (Gutierrez-pinto torrent duck, Youngflesh).

    -   Perez-Amaya (2025) - Habitat structure

-   Interactions - McQueen 2024 tested models with interactions between year × mass, year × population location, and year × migration behaviour

    -   Instead of interactions, Neate-clegg (afromontane birds) uses the full bayesian posteriors to model rates of body mass change as a function of body size

-   Include precipitation in models too as precipitation is also changing rapidly under climate change [@macedoMacroevolutionBodyExtremities2025], (Benham & Bowie, 2021 - The influence of spatially heterogeneous anthropogenic change on bill size evolution in a coastal songbird)

## Simulation

```{r}
#| label: Sim-bergs_rule_conflicting

Ex_df_hypo <- pmap(tibble(b_avg_12 = .22, r_12 = .3, r_13 = -.4, r_23 = -.1), gen_data)[[1]]

# NOTE:: In paper make note that we would know this will happen due to hypoallometry

Ex_df_hypo2 <- Ex_df_hypo %>% mutate(SA_V = Append^2 / Mass) #%>% 
  #filter(Append > 0 & Mass > 0 & SA_V < 5)
```

```{r}
#| label: Confirm-sma-slope
#| include: false

smatr::sma(Append ~ Mass, data = Ex_df_hypo2)
ref_slope <- 0.33
ref_intercept <- mean(Ex_df_hypo2$Append) - ref_slope * mean(Ex_df_hypo2$Mass)

p4 <- ggplot(Ex_df_hypo2, aes(x = Mass, y = Append)) +
  geom_point(alpha = .4) + 
  ggpmisc::stat_ma_line(method = "SMA", color = "blue") +
  geom_abline(slope = 0.33, intercept = ref_intercept, 
              linetype = "dashed", color = "red")
```

```{r}
#| label: fig-bergs_rule_conflicting

p1 <- ggplot(Ex_df_hypo2, aes(x = Temp_inc, y = Append)) +
  geom_point(alpha = .4) + 
  geom_smooth(method = "lm", se = FALSE) + 
  labs(x = "Temperature increase", y = "Appendage") 

p2 <- ggplot(Ex_df_hypo2, aes(x = Temp_inc, y = Mass)) +
  geom_point(alpha = .4) + 
  geom_smooth(method = "lm", se = FALSE) + 
  labs(x = "Temperature increase", y = "Mass")

# We would expect that SA:V would increase as temperature increases, but in this case SA:V decreases
p3 <- ggplot(Ex_df_hypo2, aes(x = Temp_inc, y = SA_V)) +
  geom_point(alpha = .4) + 
  geom_smooth(method = "lm", se = FALSE) + 
  labs(y = "Surface area to volume", x = "Temperature increase")

ggarrange(p4, p1, p2, p3, nrow = 2, ncol = 2, labels = 'AUTO')
```

Understanding the allometric scaling relationship for a species can aid in interpretation of results. For example, imagine a study using two variables as metrics of body size (wing chord and mass), and a third variable $\text{wing length}^2$ divided by body mass as a more direct metric of the surface area to volume ratio \[e.g. nwaoguTemperatureAridityDetermine2018\]. The authors are surprised as the hypothetical species appears to adhere to Bergmann's rule when mass and wing chord are used separately as the response variables, but they find a result contradicting Bergmann's rule when using $\text{wing length}^2$ divided by body mass as the response variable.

## General

The inclusion of interactions between temperature, size, and shape can reveal the co-evolution of thermoregulatory strategies in wildlife.

'Human modifications of mammal species distributions lead to substantially underestimating the strength of Bergmann’s rule (Faurby, 2016).'

There is substantial evidence documenting morphological change in response to changing climate. But what do conservation measures look like in light of our understanding of phenotypic changes? Most studies do not dare make concrete recommendations of how this might influence management or conservation planning.

Complementarity (ROODENRIJS et al black oystercatchers, 2024, SFU group)

Finally, when predicting the temperatures experienced across species’ ranges, the strongest model supported a three-way interaction between body size, tarsus length, and beak length (see @fig-interactions). Formulating the experienced climate envelope as the response variable was an especially clever way to reveal the multidimensional environment-morphology relationship.

Integrated model:

1\) Null allometric model: Mass' influence on absolute appendage length

2\) Bergmann’s rule model: The effects included in the null allometric model + temperature's influence on mass

3\) Allen’s rule model: The effects included in the null allometric model + temperature's influence on appendage length

4\) ‘Mixed’ model: The combination of models 1-3 (Allometry + Bergmann's + Allen's rules)

Still, these results were somewhat surprising given that larger birds are predicted to be more sensitive to heat stress given lower SA:V ratios (Geiser and Turbill 2009; McCain and King 2014).

Indeed, models with extensive interactions revealed codependence between size and shape in the evolutionary struggle for thermoregulation.

Indeed, @baldwinComplementarityAllensBergmanns2023 showed there was a tight negative relationship between changes in body size and appendage length, with larger shifts in one trait associated with smaller shifts in the other.

{{< pagebreak >}}

# References
